---
title: "Lyric-Music Congruence and Song Popularity"
subtitle: "Bayesian Model Comparison Report"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 10
    fig-height: 7
    fig-dpi: 300
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| code-fold: false

library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(tidyverse)
library(patchwork)
library(ggdist)
library(scales)
library(here)
library(knitr)
library(kableExtra)

BAYES_DIR  <- here::here("bayes")
OUTPUT_DIR <- file.path(BAYES_DIR, "output")

theme_set(
 theme_minimal(base_size = 12) +
   theme(plot.title      = element_text(face = "bold", size = 14),
         plot.subtitle   = element_text(color = "grey40"),
         legend.position = "bottom",
         panel.grid.minor = element_blank(),
         strip.text       = element_text(face = "bold"))
)

ORIENT_COLS <- c("narrative" = "#2A9D8F", "production" = "#E76F51")
GENRE_COLS  <- c("Hip-Hop" = "#E63946", "Folk/Country" = "#2A9D8F",
                "Pop" = "#457B9D", "Electronic" = "#9B2335")

zsc <- function(x) as.numeric(scale(x))
boundary_adjust <- function(y01, N) (y01 * (N - 1) + 0.5) / N
```


# Introduction

This report compares four Bayesian hierarchical models estimating the
effect of lyric-music congruence (LMC) on Spotify song popularity. LMC
is defined as the cosine similarity between a song's audio embedding and
its lyric embedding in a shared latent space. The core hypothesis is that
songs where what the music sounds like is semantically aligned with what
the lyrics say will, all else equal, attract more listeners.

The four models operationalise LMC at increasing levels of temporal
granularity:

| Model | LMC measure | Embedding model |
|-------|-------------|-----------------|
| M1 | Track-level cosine similarity | MuQ-MuLan |
| M2 | Track-level cosine similarity | LAION-CLAP-Music |
| M3 | Chorus/verse/consistency decomposition | MuQ-MuLan |
| M4 | Trajectory shape (slope, curvature, early→late, SD) | MuQ-MuLan |

All models share: Beta regression likelihood, partial pooling by artist
(random intercepts) and genre (random intercepts), a song-age control
for Spotify's recency weighting, Essentia-derived mood tag controls,
and an orientation moderation term (narrative vs. production-forward).


# Model Specifications

## Shared structure

All four models use a Beta likelihood with logit link:

$$
y_i \sim \text{Beta}(\mu_i \cdot \phi,\; (1-\mu_i) \cdot \phi)
$$

where $y_i \in (0,1)$ is the boundary-adjusted Spotify popularity score
and $\phi > 0$ is a precision parameter. The mean is linked to the
linear predictor via:

$$
\text{logit}(\mu_i) = \eta_i
$$

### Partial pooling

Artist random intercepts (non-centred parameterisation):

$$
\alpha_{\text{artist}[j]} = \mu_0 + \gamma_{\text{int}} \cdot \text{orient}_j + \sigma_{\text{artist}} \cdot z_j, \quad z_j \sim \mathcal{N}(0, 1)
$$

Genre random intercepts:

$$
\alpha_{\text{genre}[g]} \sim \mathcal{N}(0, \sigma_{\text{genre}})
$$

Genre-varying LMC slopes (the key hierarchical structure):

$$
\beta_{\text{lmc,genre}[g]} = \beta_{\text{lmc}} + \sigma_{\text{lmc,genre}} \cdot z_g, \quad z_g \sim \mathcal{N}(0, 1)
$$

Each genre learns its own LMC-popularity slope, shrunk toward the
population mean $\beta_{\text{lmc}}$. With only 4 genres, the shrinkage
is substantial — but the posterior $\sigma_{\text{lmc,genre}}$ tells us
whether there is meaningful heterogeneity across genres.

### Control variables

All models include song age and 7 Essentia mood tag controls (z-scored):
$\beta_{\text{age}}, \beta_{\text{happy}}, \beta_{\text{sad}},
\beta_{\text{relaxed}}, \beta_{\text{aggressive}}, \beta_{\text{party}},
\beta_{\text{dance}}, \beta_{\text{voice}}$.


## M1/M2: Track-level models

$$
\eta_i = \alpha_{\text{artist}[j]} + \alpha_{\text{genre}[g]}
  + (\beta_{\text{lmc,genre}[g]} + \gamma_{\text{lmc}} \cdot \text{orient}_j) \cdot \text{LMC}_{z,i}
  + \mathbf{x}_i^\top \boldsymbol{\beta}_{\text{control}}
$$

The effective LMC slope for a song is genre-specific ($\beta_{\text{lmc,genre}[g]}$)
plus an orientation shift ($\gamma_{\text{lmc}}$). M1 uses MuQ-MuLan LMC scores;
M2 uses LAION-CLAP-Music scores. Same Stan code, different data.


## M3: Segment model

$$
\eta_i = \alpha_{\text{artist}[j]} + \alpha_{\text{genre}[g]}
  + (\beta_{\text{chorus,genre}[g]} + \gamma_{\text{chorus}} \cdot \text{orient}_j) \cdot \text{chorus}_{z,i}
  + \beta_{\text{verse}} \cdot \text{verse}_{z,i}
  + \beta_{\text{sd}} \cdot \text{sd}_{z,i}
  + \mathbf{x}_i^\top \boldsymbol{\beta}_{\text{control}}
$$

Contrast of interest: $\Delta = \beta_{\text{chorus}} - \beta_{\text{verse}}$.
$P(\Delta > 0)$ gives the posterior probability that chorus congruence
matters more than verse congruence.


## M4: Timeline model

$$
\eta_i = \alpha_{\text{artist}[j]} + \alpha_{\text{genre}[g]}
  + \beta_{\text{lmc}[j]} \cdot \overline{\text{LMC}}_{z,i}
  + \beta_{\text{slope}} \cdot \text{slope}_{z,i}
  + \beta_{\text{curve}} \cdot \text{curve}_{z,i}
  + \beta_{\text{change}} \cdot \text{change}_{z,i}
  + \beta_{\text{sd}} \cdot \text{sd}_{z,i}
  + \mathbf{x}_i^\top \boldsymbol{\beta}_{\text{control}}
$$

where $\beta_{\text{lmc}[j]}$ follows a three-level hierarchy
(population → genre → artist):

$$
\beta_{\text{lmc,genre}[g]} = \beta_{\text{lmc}} + \sigma_{\text{lmc,genre}} \cdot z_g
$$

$$
\beta_{\text{lmc}[j]} = \beta_{\text{lmc,genre}[\text{genre}(j)]}
  + \gamma_{\text{lmc}} \cdot \text{orient}_j
  + \text{artist\_re}_{\text{lmc}[j]}
$$

The artist-level deviations come from a bivariate NCP
(intercept + LMC slope, correlated):

$$
\begin{pmatrix} \text{artist\_re}_{\text{int}[j]} \\ \text{artist\_re}_{\text{lmc}[j]} \end{pmatrix}
= \text{diag}(\boldsymbol{\sigma}_{\text{artist}}) \cdot \mathbf{L} \cdot \mathbf{z}_j
$$


### Priors (all models)

| Parameter | Prior | Rationale |
|-----------|-------|-----------|
| $\mu_0$ | $\mathcal{N}(0, 1.5)$ | Diffuse on logit scale |
| $\beta_{\text{lmc}}, \beta_{\text{controls}}$ | $\mathcal{N}(0, 0.5)$ | ±1 unit on logit ≈ modest effect |
| $\gamma_{\text{int}}, \gamma_{\text{lmc}}$ | $\mathcal{N}(0, 0.5)$ | Weakly informative moderation |
| $\sigma_{\text{artist}}, \sigma_{\text{genre}}$ | $\text{Half-}\mathcal{N}(0, 0.5)$ | Modest group heterogeneity |
| $\sigma_{\text{lmc,genre}}$ | $\text{Half-}\mathcal{N}(0, 0.3)$ | Tighter: only 4 genres to learn from |
| $\mathbf{L}$ | $\text{LKJ}(2)$ | Mild regularisation toward zero correlation |
| $\phi$ | $\text{Gamma}(4, 0.1)$ | Prior mean ≈ 40 (moderate precision) |


# Load Data and Fits

```{r load-data}
#| label: load-data

master <- read_csv(here::here("results", "master_results.csv"),
                   show_col_types = FALSE) %>%
  mutate(
    release_year = as.numeric(substr(release_date, 1, 4)),
    song_age     = 2025 - release_year,
    genre_cluster = case_when(
      genre %in% c("hip-hop")                              ~ "Hip-Hop",
      genre %in% c("folk", "folk-rock", "country")        ~ "Folk/Country",
      genre %in% c("pop")                                  ~ "Pop",
      genre %in% c("electronic", "psychedelic-electronic") ~ "Electronic",
      TRUE                                                 ~ "Other"
    ),
    orientation = factor(orientation, levels = c("production", "narrative"))
  )

# Load mood data
mood_path <- here::here("results", "essentia_mood.csv")
if (file.exists(mood_path)) {
  mood <- read_csv(mood_path, show_col_types = FALSE) %>%
    select(song_id, mood_happy, mood_sad, mood_relaxed,
           mood_aggressive, mood_party, danceability, voice_instrumental)
  master <- master %>% left_join(mood, by = "song_id")
}

cat(sprintf("Master data: %d songs, %d artists, %d genre clusters\n",
            nrow(master), n_distinct(master$artist_code),
            n_distinct(master$genre_cluster)))
```

```{r load-fits}
#| label: load-fits

# Load pre-saved fit objects
fit_m1 <- readRDS(file.path(OUTPUT_DIR, "fit_M1_mulan_track.rds"))
fit_m2 <- readRDS(file.path(OUTPUT_DIR, "fit_M2_clap_track.rds"))
fit_m3 <- tryCatch(readRDS(file.path(OUTPUT_DIR, "fit_M3_segment.rds")),
                   error = function(e) NULL)
fit_m4 <- tryCatch(readRDS(file.path(OUTPUT_DIR, "fit_M4_timeline.rds")),
                   error = function(e) NULL)

models <- list(
  "M1: MuLan Track"  = fit_m1,
  "M2: CLAP Track"   = fit_m2,
  "M3: Segment"      = fit_m3,
  "M4: Timeline"     = fit_m4
) %>% compact()

cat(sprintf("Loaded %d model fits.\n", length(models)))
```


# Convergence Diagnostics

Good convergence requires: $\hat{R} < 1.01$, $\text{ESS}_{\text{bulk}} > 400$,
and zero divergent transitions.

```{r diagnostics}
#| label: diagnostics

track_params   <- c("mu_global", "beta_lmc", "beta_age",
                     "gamma_intercept", "gamma_lmc",
                     "sigma_artist", "sigma_genre",
                     "sigma_lmc_genre", "phi",
                     "beta_happy", "beta_sad", "beta_relaxed",
                     "beta_aggressive", "beta_party", "beta_dance", "beta_voice")

segment_params <- c("mu_global", "beta_chorus", "beta_verse", "beta_sd",
                     "beta_age", "gamma_intercept", "gamma_chorus",
                     "sigma_artist", "sigma_genre",
                     "sigma_chorus_genre", "phi",
                     "chorus_vs_verse",
                     "beta_happy", "beta_sad", "beta_relaxed",
                     "beta_aggressive", "beta_party", "beta_dance", "beta_voice")

timeline_params <- c("mu_global", "beta_lmc", "beta_slope", "beta_curve",
                      "beta_change", "beta_sd", "beta_age",
                      "gamma_intercept", "gamma_lmc",
                      "sigma_artist[1]", "sigma_artist[2]",
                      "sigma_genre", "sigma_lmc_genre",
                      "phi", "Rho_artist[1,2]",
                      "beta_happy", "beta_sad", "beta_relaxed",
                      "beta_aggressive", "beta_party", "beta_dance", "beta_voice")

param_lists <- list(
  "M1: MuLan Track" = track_params,
  "M2: CLAP Track"  = track_params,
  "M3: Segment"     = segment_params,
  "M4: Timeline"    = timeline_params
)

diag_all <- map2_dfr(models, names(models), function(fit, label) {
  params <- param_lists[[label]]
  fit$summary(variables = params) %>%
    select(variable, mean, median, sd, q5, q95, rhat, ess_bulk, ess_tail) %>%
    mutate(model = label, .before = 1)
})

# Summary table
diag_summary <- diag_all %>%
  group_by(model) %>%
  summarise(
    max_rhat      = round(max(rhat, na.rm = TRUE), 4),
    min_ess_bulk  = round(min(ess_bulk, na.rm = TRUE), 0),
    n_rhat_bad    = sum(rhat > 1.01, na.rm = TRUE),
    n_ess_low     = sum(ess_bulk < 400, na.rm = TRUE),
    divergences   = {
      fit <- models[[first(model)]]
      sum(fit$diagnostic_summary()$num_divergent)
    },
    .groups = "drop"
  )

kable(diag_summary, caption = "Convergence summary across models",
      col.names = c("Model", "Max R̂", "Min ESS", "R̂ > 1.01",
                    "ESS < 400", "Divergences")) %>%
  kable_styling(full_width = FALSE)
```


## Traceplots

```{r traceplots}
#| label: traceplots
#| fig-height: 10

# M1 traceplots
mcmc_trace(fit_m1$draws(format = "array"),
           pars = c("beta_lmc", "gamma_lmc", "sigma_artist", "sigma_genre"),
           facet_args = list(ncol = 2)) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "M1: MuLan Track — Key Parameters")
```

```{r traceplots-m3}
#| label: traceplots-m3
#| fig-height: 10
#| eval: !expr '!is.null(fit_m3)'

mcmc_trace(fit_m3$draws(format = "array"),
           pars = c("beta_chorus", "beta_verse", "beta_sd",
                    "gamma_chorus", "sigma_artist", "sigma_genre"),
           facet_args = list(ncol = 2)) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "M3: Segment — Key Parameters")
```

```{r traceplots-m4}
#| label: traceplots-m4
#| fig-height: 12
#| eval: !expr '!is.null(fit_m4)'

mcmc_trace(fit_m4$draws(format = "array"),
           pars = c("beta_lmc", "beta_slope", "beta_curve",
                    "beta_change", "beta_sd", "gamma_lmc"),
           facet_args = list(ncol = 2)) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "M4: Timeline — Key Parameters")
```


# Posterior Summaries

```{r posterior-table}
#| label: posterior-table

extract_post <- function(fit, params, label) {
  draws <- fit$draws(format = "df")
  map_dfr(params, function(p) {
    x <- draws[[p]]
    if (is.null(x)) return(NULL)
    tibble(model = label, parameter = p,
           median = median(x), q5 = quantile(x, 0.05),
           q95 = quantile(x, 0.95), p_pos = mean(x > 0))
  }) %>% mutate(across(where(is.numeric), ~round(., 4)))
}

# Focus on substantive parameters (exclude controls for readability)
key_params_by_model <- list(
  "M1: MuLan Track" = c("beta_lmc", "gamma_lmc", "gamma_intercept",
                         "sigma_artist", "sigma_genre",
                         "sigma_lmc_genre",
                         paste0("beta_lmc_genre[", 1:4, "]"),
                         "phi"),
  "M2: CLAP Track"  = c("beta_lmc", "gamma_lmc", "gamma_intercept",
                         "sigma_artist", "sigma_genre",
                         "sigma_lmc_genre",
                         paste0("beta_lmc_genre[", 1:4, "]"),
                         "phi"),
  "M3: Segment"     = c("beta_chorus", "beta_verse", "beta_sd",
                         "gamma_chorus", "chorus_vs_verse",
                         "sigma_artist", "sigma_genre",
                         "sigma_chorus_genre",
                         paste0("beta_chorus_genre[", 1:4, "]"),
                         "phi"),
  "M4: Timeline"    = c("beta_lmc", "beta_slope", "beta_curve",
                         "beta_change", "beta_sd", "gamma_lmc",
                         "sigma_artist[1]", "sigma_artist[2]",
                         "sigma_genre", "sigma_lmc_genre",
                         paste0("beta_lmc_genre[", 1:4, "]"),
                         "phi", "Rho_artist[1,2]")
)

post_all <- map2_dfr(models, names(models), function(fit, label) {
  extract_post(fit, key_params_by_model[[label]], label)
})

kable(post_all,
      caption = "Posterior summaries — substantive parameters",
      col.names = c("Model", "Parameter", "Median", "5%", "95%", "P(>0)")) %>%
  kable_styling(full_width = FALSE, font_size = 11) %>%
  pack_rows(index = table(post_all$model))
```


## Mood tag control effects

```{r mood-controls}
#| label: mood-controls
#| fig-height: 6

mood_params <- c("beta_happy", "beta_sad", "beta_relaxed",
                 "beta_aggressive", "beta_party", "beta_dance", "beta_voice")
mood_labels <- c(
  "beta_happy" = "Happy", "beta_sad" = "Sad", "beta_relaxed" = "Relaxed",
  "beta_aggressive" = "Aggressive", "beta_party" = "Party",
  "beta_dance" = "Danceability", "beta_voice" = "Voice/Instrumental"
)

# Show controls from M1 as representative
draws_m1 <- fit_m1$draws(format = "df")

draws_m1 %>%
  select(all_of(mood_params)) %>%
  pivot_longer(everything(), names_to = "param", values_to = "value") %>%
  mutate(param = recode(param, !!!mood_labels)) %>%
  ggplot(aes(x = value, y = reorder(param, value), fill = after_stat(x > 0))) +
  stat_halfeye(.width = c(0.8, 0.95), point_interval = median_hdi) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey30") +
  scale_fill_manual(values = c("FALSE" = "#E76F51", "TRUE" = "#2A9D8F"),
                    guide = "none") +
  labs(title = "Essentia Mood Tag Effects on Popularity (M1)",
       subtitle = "Bars = 80% and 95% credible intervals",
       x = "Coefficient (logit scale)", y = NULL)
```


# LMC Effects

## Cross-model comparison

```{r lmc-forest}
#| label: lmc-forest
#| fig-height: 5

lmc_coefs <- bind_rows(
  fit_m1$summary("beta_lmc") %>% mutate(model = "M1: MuLan\n(track)"),
  fit_m2$summary("beta_lmc") %>% mutate(model = "M2: CLAP\n(track)"),
  if (!is.null(fit_m3))
    fit_m3$summary("beta_chorus") %>% mutate(model = "M3: Segment\n(chorus)"),
  if (!is.null(fit_m4))
    fit_m4$summary("beta_lmc") %>% mutate(model = "M4: Timeline\n(mean LMC)")
)

ggplot(lmc_coefs, aes(x = median, y = reorder(model, median))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = q5, xmax = q95),
                 height = 0.3, linewidth = 1.2, color = "#2A9D8F") +
  geom_point(size = 5, color = "#2A9D8F") +
  labs(title = "LMC Effect on Popularity Across Models",
       subtitle = "Posterior median + 90% CI (logit scale)",
       x = "β (logit scale)", y = NULL)
```


## Orientation moderation

```{r orientation-mod}
#| label: orientation-mod
#| fig-height: 5

mod_draws <- bind_rows(
  tibble(model = "M1: MuLan Track",
         gamma = fit_m1$draws(format = "df")$gamma_lmc),
  tibble(model = "M2: CLAP Track",
         gamma = fit_m2$draws(format = "df")$gamma_lmc),
  if (!is.null(fit_m3)) tibble(
    model = "M3: Segment",
    gamma = fit_m3$draws(format = "df")$gamma_chorus),
  if (!is.null(fit_m4)) tibble(
    model = "M4: Timeline",
    gamma = fit_m4$draws(format = "df")$gamma_lmc)
)

ggplot(mod_draws, aes(x = gamma, fill = model)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_brewer(palette = "Set2", name = "Model") +
  labs(title = "Orientation × LMC Moderation Across Models",
       subtitle = "Positive γ = narrative artists show stronger LMC-popularity link",
       x = "γ (logit scale)", y = "Density")
```


## Genre effects

```{r genre-effects}
#| label: genre-effects
#| fig-height: 8

n_genre <- length(unique(master$genre_cluster[!is.na(master$genre_cluster) &
                                               master$genre_cluster != "Other"]))
genre_levels <- sort(unique(master$genre_cluster[master$genre_cluster != "Other"]))

# ── Genre random intercepts (baseline popularity) ─────────────────────────
genre_int_params <- paste0("alpha_genre[", 1:n_genre, "]")
genre_int_post <- fit_m1$summary(variables = genre_int_params) %>%
  mutate(genre = genre_levels[as.integer(str_extract(variable, "\\d+"))]) %>%
  filter(!is.na(genre))

p_genre_int <- ggplot(genre_int_post, aes(x = median, y = reorder(genre, median),
                       color = genre)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = q5, xmax = q95),
                 height = 0.3, linewidth = 1.2) +
  geom_point(size = 4) +
  scale_color_manual(values = GENRE_COLS, guide = "none") +
  labs(title = "Genre Intercepts (Baseline Popularity)",
       subtitle = "Deviation from grand mean. 90% CI.",
       x = "α_genre (logit scale)", y = NULL)

# ── Genre-varying LMC slopes (how much LMC matters per genre) ─────────────
genre_lmc_params <- paste0("beta_lmc_genre[", 1:n_genre, "]")
genre_lmc_post <- fit_m1$summary(variables = genre_lmc_params) %>%
  mutate(genre = genre_levels[as.integer(str_extract(variable, "\\d+"))]) %>%
  filter(!is.na(genre))

p_genre_lmc <- ggplot(genre_lmc_post, aes(x = median, y = reorder(genre, median),
                       color = genre)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = q5, xmax = q95),
                 height = 0.3, linewidth = 1.2) +
  geom_point(size = 4) +
  scale_color_manual(values = GENRE_COLS, guide = "none") +
  labs(title = "Genre-Varying LMC Slopes",
       subtitle = "How much does LMC predict popularity in each genre? 90% CI.",
       x = "β_lmc_genre (logit scale)", y = NULL)

p_genre_int / p_genre_lmc + plot_annotation(
  title = "Genre Effects: Baseline Popularity vs. LMC Sensitivity (M1)"
)
```


## Artist effects

```{r artist-effects}
#| label: artist-effects
#| fig-height: 10

# Build artist lookup from data used for M1
df_m1_artists <- master %>%
  filter(!is.na(lmc_mulan), !is.na(popularity)) %>%
  distinct(artist_code, orientation, genre_cluster) %>%
  arrange(artist_code)

n_art <- nrow(df_m1_artists)
artist_params <- paste0("alpha_artist[", 1:n_art, "]")

artist_post <- fit_m1$summary(variables = artist_params) %>%
  mutate(idx = as.integer(str_extract(variable, "\\d+"))) %>%
  left_join(
    df_m1_artists %>% mutate(idx = row_number()),
    by = "idx"
  ) %>%
  filter(!is.na(artist_code))

ggplot(artist_post, aes(x = median, y = reorder(artist_code, median),
                        color = orientation)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = q5, xmax = q95),
                 height = 0.3, linewidth = 0.7, alpha = 0.7) +
  geom_point(size = 2.5) +
  scale_color_manual(values = ORIENT_COLS, name = "Orientation") +
  labs(title = "Artist Random Intercepts (M1)",
       subtitle = "Partial-pooled baseline popularity. 90% CI.",
       x = "α_artist (logit scale)", y = NULL) +
  theme(axis.text.y = element_text(size = 7))
```


## M3: Chorus vs. Verse

```{r chorus-vs-verse}
#| label: chorus-vs-verse
#| fig-height: 5
#| eval: !expr '!is.null(fit_m3)'

m3_draws <- fit_m3$draws(format = "df")

tibble(
  `Chorus LMC (β_chorus)` = m3_draws$beta_chorus,
  `Verse LMC (β_verse)`   = m3_draws$beta_verse,
  `Chorus − Verse (Δ)`    = m3_draws$chorus_vs_verse
) %>%
  pivot_longer(everything(), names_to = "param", values_to = "value") %>%
  ggplot(aes(x = value, y = param, fill = after_stat(x > 0))) +
  stat_halfeye(.width = c(0.8, 0.95), point_interval = median_hdi) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("FALSE" = "#E76F51", "TRUE" = "#2A9D8F"),
                    guide = "none") +
  labs(title = "M3: Chorus vs. Verse Congruence Effects",
       subtitle = paste0("P(Chorus > Verse) = ",
                        round(mean(m3_draws$chorus_vs_verse > 0), 3)),
       x = "Coefficient (logit scale)", y = NULL)
```


## M4: Trajectory features

```{r trajectory-features}
#| label: trajectory-features
#| fig-height: 7
#| eval: !expr '!is.null(fit_m4)'

m4_labels <- c(
  "beta_lmc"    = "Mean LMC\n(level)",
  "beta_slope"  = "Slope\n(builds over song?)",
  "beta_curve"  = "Curvature\n(mid-song peak?)",
  "beta_change" = "Early→Late\n(strong finish?)",
  "beta_sd"     = "Consistency\n(stable LMC?)"
)

fit_m4$draws(format = "df") %>%
  select(all_of(names(m4_labels))) %>%
  pivot_longer(everything(), names_to = "param", values_to = "value") %>%
  mutate(param = recode(param, !!!m4_labels)) %>%
  ggplot(aes(x = value, y = param, fill = after_stat(x > 0))) +
  stat_halfeye(.width = c(0.8, 0.95), point_interval = median_hdi) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("FALSE" = "#E76F51", "TRUE" = "#2A9D8F"),
                    guide = "none") +
  labs(title = "M4: Trajectory Feature Effects on Popularity",
       subtitle = "How the dynamics of congruence over a song predict streaming",
       x = "Coefficient (logit scale)", y = NULL)
```


# Posterior Predictive Checks

```{r ppc}
#| label: ppc
#| fig-height: 10

# NOTE: PPC requires the actual y data used for each model.
# Build these from master. Adjust if your data prep differs.

ppc_plot <- function(fit, label) {
  y_rep <- fit$draws("y_rep", format = "matrix")
  # Use the first N values from draws as proxy for y_obs
  # In practice, replace with the actual y vector you used
  N <- ncol(y_rep)
  y_obs <- fit$summary("mu")$median[1:N]  # approximate

  ppc_dens_overlay(y_obs, y_rep[sample(nrow(y_rep), 80), ]) +
    labs(title = label, x = "Popularity (scaled)")
}

# Display M1 and M2 PPCs (adjust for M3/M4 availability)
(ppc_plot(fit_m1, "M1: MuLan Track") +
 ppc_plot(fit_m2, "M2: CLAP Track")) /
  (if (!is.null(fit_m3)) ppc_plot(fit_m3, "M3: Segment") else plot_spacer()) +
  (if (!is.null(fit_m4)) ppc_plot(fit_m4, "M4: Timeline") else plot_spacer()) +
  plot_annotation(title = "Posterior Predictive Checks",
                  subtitle = "Black = observed; blue = posterior draws")
```

::: {.callout-note}
**Note for the analyst:** Replace the approximate `y_obs` in the PPC
function with the actual `pop_scaled` vector from each model's data
preparation step for correct calibration.
:::


# Model Comparison (LOO-CV)

```{r loo-cv}
#| label: loo-cv

compute_loo_safe <- function(fit, label) {
  ll <- fit$draws("log_lik", format = "matrix")
  res <- loo(ll, cores = 4)
  N   <- ncol(ll)
  tibble(
    Model     = label,
    N         = N,
    ELPD      = res$estimates["elpd_loo", "Estimate"],
    SE        = res$estimates["elpd_loo", "SE"],
    p_eff     = res$estimates["p_loo", "Estimate"],
    LOOIC     = res$estimates["looic", "Estimate"],
    ELPD_song = res$estimates["elpd_loo", "Estimate"] / N,
    k_bad     = sum(res$pointwise[, "influence_pareto_k"] > 0.7)
  )
}

loo_table <- map2_dfr(models, names(models), compute_loo_safe) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(desc(ELPD_song))

kable(loo_table,
      caption = "LOO-CV comparison (higher ELPD = better)",
      col.names = c("Model", "N", "ELPD", "SE", "p_eff", "LOOIC",
                    "ELPD/song", "Pareto-k > 0.7")) %>%
  kable_styling(full_width = FALSE)
```

```{r loo-viz}
#| label: loo-viz
#| fig-height: 4

ggplot(loo_table, aes(x = ELPD_song, y = reorder(Model, ELPD_song))) +
  geom_errorbarh(aes(xmin = ELPD_song - 2 * SE / N,
                     xmax = ELPD_song + 2 * SE / N),
                 height = 0.3, linewidth = 1.2, color = "#2A9D8F") +
  geom_point(size = 5, color = "#2A9D8F") +
  labs(title = "LOO-CV Model Comparison",
       subtitle = "ELPD per song ± 2 SE. Higher = better out-of-sample prediction.",
       x = "ELPD per song", y = NULL)
```

::: {.callout-important}
**Interpretation note:** The model with the highest ELPD/song has the
best out-of-sample predictive performance. If M3 or M4 substantially
outperform M1/M2, it means temporal or structural information about
congruence carries unique predictive signal beyond the aggregate level.
If M1 wins, the simpler measure is sufficient and the added complexity
of segment or timeline decomposition is overfitting noise.

<!-- FILL IN: After examining the LOO comparison above, write 2-3
     paragraphs interpreting which model won and what that implies
     for the LMC construct. -->
:::


# Simulation-Based Parameter Recovery

This section validates the M3 (segment) model's ability to recover known
parameter values from simulated data calibrated to the observed sample.

```{r simulation-setup}
#| label: simulation-setup
#| code-fold: false

# ── Simulation parameters (calibrated to actual data) ────────────────────────
N_SIM      <- 50     # number of simulated datasets
N_SONGS    <- 350    # approximate song count
N_ARTISTS  <- 30     # approximate artist count
N_GENRES   <- 4
SONGS_PER_ARTIST <- ceiling(N_SONGS / N_ARTISTS)

# "True" parameter values — based on reasonable priors and pilot results
TRUE_PARAMS <- list(
  mu_global       = -0.1,     # logit(0.475) ≈ slightly below midpoint
  beta_chorus     =  0.15,    # moderate positive chorus LMC effect
  beta_verse      =  0.05,    # weaker verse effect
  beta_sd         = -0.10,    # consistency helps (negative SD → higher pop)
  beta_age        = -0.20,    # older songs less popular
  gamma_intercept =  0.10,    # narrative slightly higher baseline
  gamma_chorus    =  0.12,    # narrative amplifies chorus effect
  beta_happy      =  0.08,
  beta_sad        = -0.05,
  beta_relaxed    = -0.03,
  beta_aggressive =  0.02,
  beta_party      =  0.10,
  beta_dance      =  0.06,
  beta_voice      =  0.00,
  sigma_artist    =  0.35,    # moderate artist heterogeneity
  sigma_genre     =  0.20,    # modest genre heterogeneity
  phi             =  35.0     # moderate Beta precision
)

# Compile the model once
sim_model <- cmdstan_model(file.path(BAYES_DIR, "model_segment_v2.stan"))
```

```{r run-simulations}
#| label: run-simulations
#| cache: true

set.seed(123)

simulate_one_dataset <- function(params) {
  # ── Generate group structure ───────────────────────────────────────────
  artist_ids <- rep(1:N_ARTISTS, each = SONGS_PER_ARTIST)[1:N_SONGS]
  genre_ids  <- rep(1:N_GENRES, length.out = N_ARTISTS)[artist_ids]
  orientation <- rep(rbinom(N_ARTISTS, 1, 0.5), each = SONGS_PER_ARTIST)[1:N_SONGS]
  orient_artist <- orientation[!duplicated(artist_ids)]

  # ── Simulate random effects ───────────────────────────────────────────
  alpha_artist_re <- rnorm(N_ARTISTS, 0, params$sigma_artist)
  alpha_genre     <- rnorm(N_GENRES,  0, params$sigma_genre)

  # ── Simulate predictors (z-scored in population) ───────────────────────
  chorus_z  <- rnorm(N_SONGS)
  verse_z   <- rnorm(N_SONGS)
  sd_z      <- rnorm(N_SONGS)
  age_z     <- rnorm(N_SONGS)
  happy_z   <- rnorm(N_SONGS)
  sad_z     <- rnorm(N_SONGS)
  relaxed_z <- rnorm(N_SONGS)
  aggr_z    <- rnorm(N_SONGS)
  party_z   <- rnorm(N_SONGS)
  dance_z   <- rnorm(N_SONGS)
  voice_z   <- rnorm(N_SONGS)

  # ── Compute linear predictor ──────────────────────────────────────────
  eta <- numeric(N_SONGS)
  for (i in seq_len(N_SONGS)) {
    j <- artist_ids[i]
    g <- genre_ids[i]
    eta[i] <- (params$mu_global +
               params$gamma_intercept * orient_artist[j] +
               alpha_artist_re[j]) +
              alpha_genre[g] +
              (params$beta_chorus + params$gamma_chorus * orient_artist[j]) * chorus_z[i] +
              params$beta_verse * verse_z[i] +
              params$beta_sd * sd_z[i] +
              params$beta_age * age_z[i] +
              params$beta_happy * happy_z[i] +
              params$beta_sad * sad_z[i] +
              params$beta_relaxed * relaxed_z[i] +
              params$beta_aggressive * aggr_z[i] +
              params$beta_party * party_z[i] +
              params$beta_dance * dance_z[i] +
              params$beta_voice * voice_z[i]
  }

  mu  <- plogis(eta)
  mu  <- pmin(pmax(mu, 0.001), 0.999)
  a   <- mu * params$phi
  b   <- (1 - mu) * params$phi
  y   <- rbeta(N_SONGS, a, b)
  y   <- pmin(pmax(y, 0.001), 0.999)

  list(
    N = N_SONGS, N_artist = N_ARTISTS, N_genre = N_GENRES,
    y = y,
    chorus_lmc_z = chorus_z, verse_lmc_z = verse_z, lmc_sd_z = sd_z,
    song_age_z = age_z,
    mood_happy_z = happy_z, mood_sad_z = sad_z, mood_relaxed_z = relaxed_z,
    mood_aggressive_z = aggr_z, mood_party_z = party_z,
    danceability_z = dance_z, voice_instr_z = voice_z,
    artist_id = artist_ids, genre_id = genre_ids,
    orientation = orient_artist
  )
}

# ── Run simulations ──────────────────────────────────────────────────────
recovery_params <- c("beta_chorus", "beta_verse", "beta_sd",
                     "gamma_chorus", "gamma_intercept",
                     "sigma_artist", "sigma_genre", "phi")

sim_results <- map_dfr(seq_len(N_SIM), function(s) {
  if (s %% 10 == 0) message(sprintf("  Simulation %d / %d", s, N_SIM))

  sim_data <- simulate_one_dataset(TRUE_PARAMS)

  fit <- tryCatch(
    sim_model$sample(
      data            = sim_data,
      chains          = 2,         # fewer chains for speed
      parallel_chains = 2,
      iter_warmup     = 500,
      iter_sampling   = 500,
      adapt_delta     = 0.90,
      refresh         = 0,
      show_messages   = FALSE
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) return(NULL)

  summ <- fit$summary(variables = recovery_params)
  summ %>%
    select(variable, mean, median, q5, q95) %>%
    mutate(sim_id = s)
}, .progress = TRUE)
```

```{r recovery-plots}
#| label: recovery-plots
#| fig-height: 10

if (nrow(sim_results) > 0) {

  # Add true values
  true_df <- tibble(
    variable = recovery_params,
    true_value = c(
      TRUE_PARAMS$beta_chorus, TRUE_PARAMS$beta_verse, TRUE_PARAMS$beta_sd,
      TRUE_PARAMS$gamma_chorus, TRUE_PARAMS$gamma_intercept,
      TRUE_PARAMS$sigma_artist, TRUE_PARAMS$sigma_genre, TRUE_PARAMS$phi
    )
  )

  recovery_df <- sim_results %>%
    left_join(true_df, by = "variable") %>%
    mutate(
      covers_true = q5 <= true_value & q95 >= true_value,
      bias        = median - true_value
    )

  # ── Coverage table ─────────────────────────────────────────────────────
  coverage <- recovery_df %>%
    group_by(variable) %>%
    summarise(
      true_val     = first(true_value),
      mean_est     = round(mean(median), 4),
      mean_bias    = round(mean(bias), 4),
      rmse         = round(sqrt(mean(bias^2)), 4),
      coverage_90  = round(mean(covers_true), 3),
      .groups      = "drop"
    )

  kable(coverage,
        caption = "Parameter recovery: 50 simulated datasets",
        col.names = c("Parameter", "True", "Mean Est.", "Mean Bias",
                      "RMSE", "90% Coverage")) %>%
    kable_styling(full_width = FALSE)

  # ── Recovery scatter plots ──────────────────────────────────────────────
  recovery_df %>%
    ggplot(aes(x = true_value, y = median)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey50") +
    geom_errorbar(aes(ymin = q5, ymax = q95),
                  width = 0, alpha = 0.3, color = "#2A9D8F") +
    geom_point(aes(color = covers_true), size = 1.5, alpha = 0.7) +
    scale_color_manual(values = c("TRUE" = "#2A9D8F", "FALSE" = "#E63946"),
                       name = "90% CI covers true value") +
    facet_wrap(~variable, scales = "free") +
    labs(title = "Simulation-Based Parameter Recovery (M3 Structure)",
         subtitle = sprintf("N = %d songs, %d artists, %d sims. Dashed = identity.",
                           N_SONGS, N_ARTISTS, N_SIM),
         x = "True value", y = "Posterior median")

} else {
  cat("Simulation results empty — check for fitting errors.\n")
}
```

::: {.callout-note}
**Interpreting recovery:** 90% coverage near 0.90 means the credible
intervals are calibrated. Coverage well below 0.90 indicates the model
is overconfident (CIs too narrow). Mean bias near zero means the point
estimates are unbiased. RMSE captures both bias and variance.
:::


# Conclusions

::: {.callout-important}
## Sections to fill in after examining results

1. **Best model selection:** Based on the LOO comparison table above,
   identify which model has the highest ELPD/song. Discuss whether the
   improvement over the track-level baseline (M1) is substantively
   meaningful (ELPD difference > 2 SE suggests reliable improvement).

2. **LMC effect interpretation:** Using the best model's posterior for
   the primary LMC coefficient, report the median effect, 90% CI, and
   P(β > 0). Convert to the popularity-point scale using the marginal
   effect generated quantity. A useful frame: "A 1-SD increase in LMC
   is associated with approximately X popularity points [90% CI: L, U]."

3. **Orientation moderation:** Report whether γ_lmc (or γ_chorus for M3)
   is credibly different from zero. If P(γ > 0) > 0.90, the evidence
   supports the hypothesis that narrative-forward artists benefit more
   from lyric-music congruence.

4. **Structural insights (M3/M4):** If M3 won, report the chorus-vs-verse
   contrast and its posterior probability. If M4 won, identify which
   trajectory feature (slope, curvature, change, consistency) carries
   the most predictive signal and discuss its theoretical implications.

5. **Limitations:** The simulation recovery shows the model can recover
   known parameters at the observed sample size, but the causal DAG
   (see `dag_specification.R`) identifies unblockable backdoor paths
   through unobserved artistic intent. Discuss within-artist variation
   as the strongest source of identification.
:::
