---
title: "Lyric–Music Congruence and Song Popularity"
subtitle: "A Working Paper on the Cross-Modal Drivers of Streaming Engagement"
author:
  - name: Ethan Budge
    affiliation: |
      Ph.D. Student, Department of Marketing and Logistics  
      Fisher College of Business, The Ohio State University
    email: budge.13@osu.edu
date: today
abstract: |
  This working paper introduces *lyric–music congruence* (LMC) — the
  cosine similarity between audio and lyric embeddings in a
  jointly-trained affective space — as a candidate driver of song
  popularity on streaming platforms. Drawing on processing fluency
  theory,narrative transportation, and the current research on joint 
  embedding spaces, I hypothesize that LMC raises Spotify popularity, 
  with stronger effects for narrative-forward
  artists. I report preliminary Bayesian estimates from four
  hierarchical specifications fit to a corpus of 359 songs across 34
  artists and four genre clusters. The population-level LMC effect is
  statistically inconclusive at the current sample, with 90% credible
  intervals straddling zero in every specification. Three structural
  features of LMC carry directional but sub-decisive signal in the
  predicted direction: chorus congruence outweighs verse congruence
  (P[Δ > 0] = 0.64), within-song consistency of LMC modestly predicts
  higher popularity (P[β < 0] = 0.96), and the orientation moderation
  is positive in the trajectory model (P[γ > 0] = 0.75). The
  segment-level model provides the best out-of-sample fit. I treat
  these findings as a first pass that has established the measurement
  pipeline, identified design choices, and quantified the data and
  identification challenges that remain. The pipeline generalizes to
  other multimodal marketing contexts including advertising
  music–visual alignment and brand narrative coherence.
keywords:
  - multimodal emotion recognition
  - lyric–music congruence
  - music marketing
  - processing fluency
  - narrative transportation
  - streaming popularity
  - contrastive learning
  - audio–text embeddings
format:
  pdf:
    documentclass: article
    classoption:
      - 11pt
    geometry:
      - margin=1in
    fig-pos: "H"
    fig-cap-location: bottom
    tbl-cap-location: top
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{setspace}
        \doublespacing
        \usepackage{booktabs}
        \usepackage{amsmath}
        \usepackage{amssymb}
        \usepackage{mathtools}
        \usepackage{microtype}
        \setlength{\parindent}{0.4in}
number-sections: true
toc: false
execute:
  warning: false
  message: false
  echo: false
---

```{r setup}
#| include: false

library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(tidyverse)
library(patchwork)
library(ggdist)
library(scales)
library(here)
library(knitr)
library(kableExtra)

BAYES_DIR  <- here::here("bayes")
OUTPUT_DIR <- file.path(BAYES_DIR, "output")

theme_set(
  theme_minimal(base_size = 11) +
    theme(plot.title      = element_text(face = "bold", size = 12),
          plot.subtitle   = element_text(color = "grey40"),
          legend.position = "bottom",
          panel.grid.minor = element_blank(),
          strip.text       = element_text(face = "bold"))
)

ORIENT_COLS <- c("narrative" = "#2A9D8F", "production" = "#E76F51")
GENRE_COLS  <- c("Hip-Hop" = "#E63946", "Folk/Country" = "#2A9D8F",
                 "Pop" = "#457B9D", "Electronic" = "#9B2335")

zsc <- function(x) as.numeric(scale(x))
boundary_adjust <- function(y01, N) (y01 * (N - 1) + 0.5) / N
```

# Introduction

A song can communicate emotions and ideas through both its acoustic properties
and its lyrical content. Listeners experience both at once, and the
relationship between the two is itself a source of meaning, pleasure,
or dissonance. Yet quantitative marketing research has largely
treated the audio and lyrical dimensions of music as separate objects
of study. This paper proposes and begins to empirically test a
unified measure of *lyric–music congruence* (LMC) — the degree to
which a song's acoustic emotional signal and its lyrical emotional
signal occupy similar regions of a shared affective embedding space —
and examines its consequences for commercial success on streaming
platforms.

The impetus is threefold. First, the economic stakes are high. Global
recorded music revenues reached \$28.6 billion in 2023, with
streaming accounting for over 67% of that total (IFPI 2024); on
Spotify alone, more than 100,000 tracks are uploaded daily, creating
an attention economy in which fine-grained design differences carry
outsized commercial consequences. Second, recent advances in
multimodal machine learning — contrastive representation learning in
the tradition of CLIP (Radford et al. 2021), the audio–text model
CLAP (Wu et al. 2023), and Google's music–language joint embedding
MuLan (Huang et al. 2022) — now make it feasible to encode raw audio
and natural-language lyrics into a shared latent space, enabling a
model-agnostic measure of emotional alignment that requires no a
priori commitment to a discrete emotion taxonomy. Third, although both audio and lyrics have been extensively studied from psychological and qualitative perspectives (SOURCE), the cross-modal relationship between them has not been quantitatively operationalized or linked to consumer behavior at scale before.

This paper makes three intended contributions. **Methodologically**,
I introduce LMC as a construct and develop a pipeline for computing
it at scale using pre-trained audio and lyric encoders embedded in a
contrastively-aligned joint space; the pipeline generalizes to other
multimodal marketing contexts. **Empirically**, I estimate the
effect of LMC on Spotify's recency-weighted popularity score using
four Bayesian hierarchical specifications that vary in temporal
granularity. **Theoretically**, I hypothesize heterogeneity in the
mechanisms linking LMC to popularity: narrative-forward songs are
predicted to operate primarily through transportation, and
production-forward songs through fluency. This is a substantial potential finding in and of itself while also opening routes for future quantitative research on abstract psychological mechanisms.

As the work currently stands, the
pipeline is built and operational, but the empirical analysis is at
an early stage. My corpus of 359 songs across 34 artists yields
adequate computational diagnostics but is not large enough to deliver
decisive evidence. The population-level LMC coefficient in my
preferred specification has a posterior median of $0.006$ on the
logit scale with a 90% credible interval of $[-0.131, 0.131]$,
leaving its sign and magnitude effectively undetermined. However, as a current state of progress, I report what the data suggest at directional,
sub-decisive levels: chorus-level LMC outperforms verse-level LMC,
within-song consistency predicts higher popularity, and the
moderation by orientation is in the predicted direction in the
trajectory model. These signals motivate the next phase, laid out at
the end of the paper.

The paper proceeds as follows. Section \ref{literature-review}
reviews the relevant literature. Section
\ref{conceptual-framework-and-hypotheses} develops the conceptual
framework and states formal hypotheses. Section
\ref{data-and-measurement} describes the data and embedding pipeline.
Section \ref{empirical-strategy} presents the hierarchical Bayesian
specification. Section \ref{results} reports results. Section
\ref{discussion} discusses what the current evidence does and does
not allow us to conclude. Section
\ref{limitations-and-the-path-forward} lays out the path forward.

# Literature Review

## Emotion in Music: Psychoacoustic Foundations and Consumer Response

A foundational distinction in the music psychology literature is
between *perceived* emotion — what a listener recognizes as the
emotional character of a piece — and *induced* emotion — what the
listener actually experiences (Eerola and Vuoskoski 2011; Gabrielsson
2002). Eerola and Vuoskoski (2011) compared seven major emotion models
applied to identical stimuli and found that no single framework —
discrete, dimensional, or prototype-based — provided uniformly
superior classification accuracy, motivating taxonomy-free approaches
to emotional representation.

In the consumer behavior literature, cross-domain alignment of music
with consumption environments has been shown to matter. North,
Hargreaves, and McKendrick (1999) showed in a wine shop field
experiment that nationally congruent background music influenced
product choice; Areni and Kim (1993) showed that classical versus
Top-40 background music shifted consumers toward higher-priced wine
selections; and MacInnis and Park (1991) distinguished between the
"fit" of music to ad content and music's arousal properties as
independent predictors of ad attitude. These studies established
cross-modal coherence as a consumer driver but predated the
computational tools needed to measure it at scale.

## Computational Music Emotion Recognition

Computational music emotion recognition has progressed from the previous standard of 
hand-crafted acoustic features mapped onto dimensional emotion models, inferring emotions from features such as tempo, mode, spectral centroid, and MFCCs (Yang and Chen 2012). Panda,
Malheiro, and Paiva (2023) document that arousal is substantially
easier to predict from audio than valence. A central limitation is
dependence on predefined emotion taxonomies that may not correspond
to the continuous, high-dimensional nature of musical emotion. Fong,
Kumar, and Sudhir (2024) recently developed a theory-based explainable
CNN whose harmonics-grounded filters achieve competitive prediction
while preserving interpretability — demonstrating that audio
embeddings carrying emotional information can be extracted in ways
tractable for marketing-style downstream models. A separate strand
has built jointly-trained audio–text models: MuLan (Huang et al.
2022) over 44 million music recordings paired with playlist
annotations, and LAION-CLAP (Wu et al. 2023) over a general-purpose
caption corpus. These models do not require predefined emotion
labels; the alignment of audio and text within their learned spaces
becomes itself a measurable quantity, and this is what makes a
continuous, scale-free measure of LMC feasible.

## Processing Fluency and Aesthetic Preference

A robust finding across the cognitive and consumer literatures is that
stimuli which are easier to process are evaluated more favorably
(Reber, Schwarz, and Winkielman 2004; Winkielman et al. 2006).
Processing fluency generates affect that is misattributed to the
stimulus, producing aesthetic preference. Landwehr, Labroo, and
Herrmann (2011) showed that design fluency measures improve automobile
sales forecasts, and Hekkert (2006) integrated fluency into a broader
inverted-U model of design aesthetics. Cross-modally, when two
channels convey congruent information, the listener does not have to
reconcile competing representations and fluency is enhanced (Lee and
Labroo 2004). A song whose musical affect aligns with its lyrical
affect imposes lower processing demands than one whose channels
diverge, and should therefore be preferred. Melvill-Smith, Krause, and
North (2023) applied processing fluency directly to lyrics, finding
that lyric simplicity — operationalized through readability and
repetitiveness — predicted UK chart success between 1999 and 2013.
Their analysis is confined to the lyric channel alone and does not
consider whether the alignment of lyric and musical affect constitutes
an independent driver of popularity.

## Narrative Transportation and Lyrical Engagement

For songs in which lyrical content functions narratively — inviting
the listener to follow a story, inhabit a perspective, or trace an
emotional arc — a distinct mechanism may operate. Green and Brock
(2000) define *transportation* as a state of absorption in a narrative
in which cognitive and affective resources are directed toward the
story world. Transported listeners show stronger affective response,
more vivid imagery, and stronger belief change; transportation is
disrupted by "false notes" — narrative incoherence or affective
mismatch — that break immersion. Van Laer et al. (2014) meta-analyzed
76 studies and documented significant effects on persuasion and
attitude change. In advertising, Phillips and McQuarrie (2010) found
that emotionally congruent visual–verbal combinations produced
stronger narrative engagement and more favorable brand attitudes than
incongruent ones — a paradigm structurally parallel to music–lyric
relations. The music constitutes the affective "world" into which the
listener is invited; the lyrics are the narrative that unfolds within
it. When emotional character and emotional register align,
transportation is facilitated. When they diverge, the listener faces
what I term *affective narrative friction*.

The transportation account generates predictions distinct from
fluency and implies a moderating role for lyrical orientation. For
narrative-forward songs, congruence operates by sustaining immersion;
for production-forward songs, through cognitive ease. Both pathways
predict a positive LMC effect, but they differ in *who benefits most*.

## Music Marketing and Streaming Popularity

The marketing literature on music popularity has historically focused
on chart outcomes and acoustic features. Salganik, Dodds, and Watts
(2006) demonstrated experimentally that musical success is far more
unpredictable than fundamentals alone would suggest, with social
influence amplifying small initial differences. Askin and Mauskapf
(2017) found an inverted-U between acoustic typicality and Billboard
chart position, implying that the relationship between any coherence
measure and popularity may itself be non-linear. Ordanini, Nunes, and
Nanni (2018) found that genre-distant collaborations achieved greater
chart success, consistent with the co-branding literature.

We adopt Spotify's proprietary popularity score — a platform-computed
index from 0 to 100 reflecting stream volume and recency, with recent
activity weighted more heavily (Spotify AB 2024). The recency
weighting rewards sustained engagement over promotional spikes. We
note that streaming popularity is a coarse outcome variable, not
separately identifying discovery, retention, repeat listening, or
skip behavior — each of which would be more diagnostic of the
mechanisms LMC is hypothesized to engage. Ideally, this program would
obtain listener-level behavioral data through a partnership with
Spotify, Apple Music, or another platform. Such a partnership has
not yet materialized; in its absence I proceed with the
publicly-available popularity index scraped via the Spotify Web API,
flagging this as a substantive limitation I aim to address.

## Genre Heterogeneity and the Production–Narrative Spectrum

Genre represents not only musical conventions but distinct listener
populations and the functional role of lyrics relative to
instrumentation (Lena and Peterson 2008). In hip-hop and country,
lyrical craft and narrative authenticity carry high cultural value;
in EDM and mainstream pop, minimal or textural lyrics are frequently
subordinated to sonic production. The emotional significance of any
mismatch between lyrical and musical channels therefore varies
systematically across genre contexts. The "text as data" literature
in marketing has demonstrated that high-dimensional text
representations contain commercially relevant signals poorly
approximated by hand-coded categories (Berger et al. 2020), with
Toubia et al. (2019) using embedding-derived ideation diversity to
predict new product performance and Dew and Ansari (2018) applying
Bayesian nonparametric models to music preference heterogeneity.

Most prior music emotion recognition work has treated audio and
lyrics as parallel channels to be fused for prediction of a common
emotion label, rather than as channels whose mutual alignment might
itself have behavioral consequences. The music marketing literature
has studied acoustic features and chart outcomes, or lyrical
complexity and popularity, but not the cross-modal relationship
between them. This intersection is the gap this research is designed
to fill.

# Conceptual Framework and Hypotheses

The assumed data-generating process can be summarized as follows. Three
upstream factors — artist identity, genre, and lyric orientation —
jointly shape both the song's audio features and its lyrical content.
The resulting song produces an audio recording and a lyric stream
whose joint embedding gives rise to the LMC measure that is the focus
of inference. Popularity is downstream of LMC and a set of unobserved
platform, advertising, and social distribution forces; the path from
LMC to popularity is identified up to bias from these forces, and we
return to the implications in Section \ref{discussion}. A diagrammatic
version of this DAG appears in the companion presentation.

The construct of central interest is *lyric–music congruence*, defined
as the cosine similarity between a song's audio embedding and its
lyric embedding in a jointly-trained high-dimensional space. We
operationalize LMC at four levels of granularity and treat orientation
as a binary classification of artist into *narrative-forward* versus
*production-forward* based on genre conventions. Four formal
hypotheses follow.

**Hypothesis 1 (Main effect).** *Holding artist, genre, song age, and
audio mood fixed, higher LMC predicts higher Spotify popularity.*
Both fluency and transportation predict a positive coefficient; the
two accounts differ on mechanism, not direction.

**Hypothesis 2 (Orientation moderation).** *The positive effect of LMC
on popularity is stronger for narrative-forward artists than for
production-forward artists.* This follows from transportation: when
lyrics carry the cognitive and affective load, congruence with the
musical world has more to disrupt or sustain.

**Hypothesis 3a (Chorus over verse).** *Chorus-level LMC has a larger
positive effect on popularity than verse-level LMC.* Choruses are
repeated, hooked, and disproportionately recalled.

**Hypothesis 3b (Within-song consistency).** *Songs with greater
within-song consistency of LMC — low standard deviation of segment-level
LMC — have higher popularity, all else equal.* Inconsistent congruence
creates moments of disrupted fluency or narrative friction.

**Hypothesis 4 (Trajectory shape).** *Beyond mean LMC, the dynamics of
LMC over a song carry incremental signal: positive slope and modest
mid-song peaks (positive curvature) predict higher popularity.* This
hypothesis is exploratory.

# Data and Measurement

## Sampling Frame

The current corpus consists of 34 artists distributed across four
genre clusters: hip-hop (10 artists), folk/Americana/country (9
artists), pop (8 artists), and electronic (7 artists). The first two
clusters are operationalized as *narrative-forward*, the second two
as *production-forward*, following the literature on genre conventions
in lyrical function (Lena and Peterson 2008). Within each artist, we
collected approximately 10–15 representative tracks spanning the
discography, biased toward songs with meaningful streaming history.
We sampled 441 candidate songs; due to incomplete pairings of audio
with lyrics, the analytic corpus retained for the track-level models
is 359 songs. The subset retained for the segment model (M3) is 324
songs; the subset for the line-by-line trajectory model (M4) is 114
songs. Sparsity in the higher-granularity models reflects the
difficulty of automatically matching lyric segments to audio
timestamps for a portion of the corpus.

We acknowledge that this is not a probability sample of all music on
streaming platforms. Artists were purposively selected to cover the
four-cluster design, and within-artist song selection biases toward
better-known tracks. The corpus therefore over-represents
already-popular songs. Estimates of LMC slopes should be interpreted
within the realized sample; the population to which they generalize
is *commercially active artists in the four sampled genre clusters*,
not music on Spotify in general. Section
\ref{limitations-and-the-path-forward} describes the systematic
sampling extension I plan.

## Pipeline

The data pipeline consists of five stages. (1) **Lyric scraping**
queries Genius via the LyricsGenius API and persists raw lyric text
plus structural section headers (`[Verse 1]`, `[Chorus]`, etc.).
(2) **Audio acquisition** retrieves audio for each song via `yt-dlp`
and converts to WAV at the sample rate required by each downstream
encoder. (3) **Schema construction** validates the cross-product of
audio and lyric availability into a unified catalog. (4) **Embedding**
runs each of three encoder configurations (described below) over the
full corpus. (5) **Segment analysis** uses WhisperX (Bain et al. 2023)
for forced alignment of lyric lines to audio timestamps, falling back
to equal-duration splits when alignment confidence is insufficient.
Spotify popularity scores, release dates, and Essentia mood controls
are joined at the song level for the analytic file.

## Embedding Models

We compute LMC under three encoder configurations to characterize
sensitivity of results to embedding choice.

**MuQ-MuLan** (Huang et al. 2022) is a music–language joint embedding
model trained on 44 million music recordings paired with playlist
tags. Audio is embedded at 24 kHz and lyrics as a single text input;
both vectors live in the same 512-dimensional space, making cosine
similarity directly interpretable as semantic alignment. This is the
primary encoder for my analyses.

**LAION-CLAP-Music** (Wu et al. 2023) is a contrastive language–audio
pretraining model with broader training data (630K audio–caption
pairs spanning music, environmental sounds, and AudioSet). Its audio
representations carry stronger perceptual and timbral information
relative to MuLan's tag-driven training. Comparing CLAP- and
MuLan-derived LMC tests robustness to encoder choice.

## LMC Operationalizations

We construct four operationalizations giving rise to the four models
in Section \ref{empirical-strategy}.

**Track-level LMC** is the cosine similarity between the full-track
audio and full-lyric embeddings — computationally efficient but blind
to internal structure. I compute this under both MuLan (M1) and
CLAP (M2).

**Segment-level LMC** decomposes each song into chorus and verse
segments using the structural headers preserved during scraping. The
corresponding audio interval is identified via WhisperX forced
alignment, embedded separately, and cosine similarity to the segment
lyric embedding is computed. I aggregate to song-level summaries:
mean chorus LMC, mean verse LMC, and within-song standard deviation
(a consistency measure). This is the input to M3.

**Line-level LMC** decomposes each song line by line, computing LMC
for each lyric line against its temporally-aligned audio interval.
From the time series I extract four trajectory features — mean,
slope, curvature, and early-to-late change — plus the standard
deviation. This is the input to M4. The line-level model has the
smallest analytic sample (N = 114) because line-by-line forced
alignment fails for a large share of the corpus.

## Outcome and Controls

Spotify popularity is retrieved via the Spotify Web API; the score is
bounded $[0, 100]$, and I rescale to the open unit interval $(0, 1)$
via boundary adjustment $y' = (y(N-1) + 0.5)/N$ (Smithson and
Verkuilen 2006). *Song age* (years since release) absorbs the
mechanical penalty Spotify's recency weighting imposes on older songs.
A vector of seven *Essentia-derived mood controls* (happy, sad,
relaxed, aggressive, party, danceability, voice/instrumental balance)
from pretrained classifiers (Bogdanov et al. 2013) is z-standardized
and entered into the linear predictor; these absorb variation in the
audio's affective character independent of its lyric alignment, so
the LMC coefficient identifies cross-modal alignment rather than
overall mood. Unobserved drivers — promotional intensity, playlist
placement, social media virality, listener demographics — remain in
the error term; I take up the implications in Section
\ref{discussion}.

# Empirical Strategy

```{r load-data}
#| include: false

master_path <- here::here("results", "master_results.csv")

if (file.exists(master_path)) {
  master <- read_csv(master_path, show_col_types = FALSE) %>%
    mutate(
      release_year = as.numeric(substr(release_date, 1, 4)),
      song_age     = 2025 - release_year,
      genre_cluster = case_when(
        genre %in% c("hip-hop")                              ~ "Hip-Hop",
        genre %in% c("folk", "folk-rock", "country")         ~ "Folk/Country",
        genre %in% c("pop")                                  ~ "Pop",
        genre %in% c("electronic", "psychedelic-electronic") ~ "Electronic",
        TRUE                                                 ~ "Other"
      ),
      orientation = factor(orientation, levels = c("production", "narrative"))
    )

  mood_path <- here::here("results", "essentia_mood.csv")
  if (file.exists(mood_path)) {
    mood <- read_csv(mood_path, show_col_types = FALSE) %>%
      select(song_id, mood_happy, mood_sad, mood_relaxed,
             mood_aggressive, mood_party, danceability, voice_instrumental)
    master <- master %>% left_join(mood, by = "song_id")
  }
} else {
  master <- NULL
}
```

```{r load-fits}
#| include: false

safe_read <- function(p) tryCatch(readRDS(p), error = function(e) NULL)

fit_m1 <- safe_read(file.path(OUTPUT_DIR, "fit_M1_mulan_track.rds"))
fit_m2 <- safe_read(file.path(OUTPUT_DIR, "fit_M2_clap_track.rds"))
fit_m3 <- safe_read(file.path(OUTPUT_DIR, "fit_M3_segment.rds"))
fit_m4 <- safe_read(file.path(OUTPUT_DIR, "fit_M4_timeline.rds"))

models <- list(
  "M1: MuLan Track"  = fit_m1,
  "M2: CLAP Track"   = fit_m2,
  "M3: Segment"      = fit_m3,
  "M4: Timeline"     = fit_m4
) %>% compact()
```

## Beta Regression with Hierarchical Partial Pooling

All four models share a common likelihood. Let $y_i \in (0, 1)$ denote
the boundary-adjusted Spotify popularity for song $i$. I assume

$$
y_i \sim \text{Beta}(\mu_i \cdot \phi,\; (1 - \mu_i) \cdot \phi),
$$

\noindent where $\phi > 0$ is a precision parameter and the mean is
linked to the linear predictor by $\text{logit}(\mu_i) = \eta_i$. Beta
regression is natural given bounded outcome support (Ferrari and
Cribari-Neto 2004).

The linear predictor incorporates partial pooling at two levels.
Artists are nested within four genre clusters; I estimate random
intercepts at both levels. The artist intercept uses a non-centered
parameterization to improve sampling geometry:

$$
\alpha_{\text{artist}[j]} = \mu_0 + \gamma_{\text{int}} \cdot \text{orient}_j + \sigma_{\text{artist}} \cdot z_j, \quad z_j \sim \mathcal{N}(0, 1),
$$

\noindent where $\mu_0$ is the grand mean and $\gamma_{\text{int}}$ a
fixed orientation shift. Genre intercepts are
$\alpha_{\text{genre}[g]} \sim \mathcal{N}(0, \sigma_{\text{genre}})$.
We allow the LMC slope to vary by genre:

$$
\beta_{\text{lmc, genre}[g]} = \beta_{\text{lmc}} + \sigma_{\text{lmc, genre}} \cdot z_g, \quad z_g \sim \mathcal{N}(0, 1).
$$

\noindent The posterior on $\sigma_{\text{lmc, genre}}$ tests for
heterogeneity in LMC sensitivity across genres. With only four genres,
partial pooling shrinks aggressive estimates toward the population
mean. All models include song age and seven mood controls, denoted
collectively $\mathbf{x}_i^\top \boldsymbol{\beta}_{\text{control}}$.

## The Four Models

**Model M1 (MuLan track-level)**:

$$
\eta_i^{M1} = \alpha_{\text{artist}[j]} + \alpha_{\text{genre}[g]} + (\beta_{\text{lmc, genre}[g]} + \gamma_{\text{lmc}} \cdot \text{orient}_j) \cdot \text{LMC}_{z, i}^{\text{MuLan}} + \mathbf{x}_i^\top \boldsymbol{\beta}_{\text{control}}
$$

\noindent The orientation shift $\gamma_{\text{lmc}}$ tests Hypothesis 2:
$\gamma_{\text{lmc}} > 0$ implies narrative-forward artists benefit
more from LMC than production-forward artists.

**Model M2 (CLAP track-level)** is identical to M1 with $\text{LMC}^{\text{CLAP}}$
in place of $\text{LMC}^{\text{MuLan}}$. Comparing M1 and M2
characterizes encoder sensitivity.

**Model M3 (segment)** decomposes track LMC into chorus and verse
components and adds a within-song consistency term:

$$
\eta_i^{M3} = \alpha_{\text{artist}[j]} + \alpha_{\text{genre}[g]} + (\beta_{\text{chorus, genre}[g]} + \gamma_{\text{chorus}} \cdot \text{orient}_j) \cdot \text{chorus}_{z, i} + \beta_{\text{verse}} \cdot \text{verse}_{z, i} + \beta_{\text{sd}} \cdot \text{sd}_{z, i} + \mathbf{x}_i^\top \boldsymbol{\beta}_{\text{control}}
$$

\noindent The contrast $\Delta = \beta_{\text{chorus}} -
\beta_{\text{verse}}$ tests Hypothesis 3a; $\beta_{\text{sd}}$ tests
Hypothesis 3b.

**Model M4 (trajectory)** uses the line-by-line LMC time series and
extracts four trajectory features in addition to mean LMC:

$$
\eta_i^{M4} = \alpha_{\text{artist}[j]} + \alpha_{\text{genre}[g]} + \beta_{\text{lmc}[j]} \cdot \overline{\text{LMC}}_{z, i} + \beta_{\text{slope}} \cdot \text{slope}_{z, i} + \beta_{\text{curve}} \cdot \text{curve}_{z, i} + \beta_{\text{change}} \cdot \text{change}_{z, i} + \beta_{\text{sd}} \cdot \text{sd}_{z, i} + \mathbf{x}_i^\top \boldsymbol{\beta}_{\text{control}}
$$

\noindent where $\beta_{\text{lmc}[j]}$ follows a three-level hierarchy
(population $\to$ genre $\to$ artist) and artist-level deviations come
from a bivariate non-centered parameterization on intercept and LMC
slope with LKJ-prior correlation:

$$
\begin{pmatrix} \text{artist\_re}_{\text{int}[j]} \\ \text{artist\_re}_{\text{lmc}[j]} \end{pmatrix} = \text{diag}(\boldsymbol{\sigma}_{\text{artist}}) \cdot \mathbf{L} \cdot \mathbf{z}_j.
$$

## Priors

We use weakly informative priors throughout, summarized in Table
\ref{tab:priors}. The grand mean $\mu_0$ has a diffuse prior on the
logit scale; slope parameters have $\mathcal{N}(0, 0.5)$ priors,
implying a one-unit logit-scale shift is plausible but not preferred.
Variance components use half-normal priors. The LKJ(2) prior mildly
regularizes correlations toward zero. The precision $\phi$ has a
Gamma(4, 0.1) prior with mean 40, consistent with moderate Beta
dispersion.

\begin{table}[H]
\centering
\caption{Priors used across all four models.\label{tab:priors}}
\begin{tabular}{lll}
\toprule
\textbf{Parameter} & \textbf{Prior} & \textbf{Rationale} \\
\midrule
$\mu_0$ & $\mathcal{N}(0, 1.5)$ & Diffuse on logit scale \\
$\beta_{\text{lmc}}, \boldsymbol{\beta}_{\text{control}}$ & $\mathcal{N}(0, 0.5)$ & ±1 unit on logit $\approx$ modest effect \\
$\gamma_{\text{int}}, \gamma_{\text{lmc}}$ & $\mathcal{N}(0, 0.5)$ & Weakly informative moderation \\
$\sigma_{\text{artist}}, \sigma_{\text{genre}}$ & Half-$\mathcal{N}(0, 0.5)$ & Modest group heterogeneity \\
$\sigma_{\text{lmc, genre}}$ & Half-$\mathcal{N}(0, 0.3)$ & Tighter: only 4 genres \\
$\mathbf{L}$ & LKJ(2) & Mild regularization to zero correlation \\
$\phi$ & Gamma(4, 0.1) & Prior mean $\approx 40$ \\
\bottomrule
\end{tabular}
\end{table}

## Estimation

All four models are fit in cmdstanr (Stan Development Team 2024) with
four chains of 4,000 iterations (2,000 warmup, 2,000 sampling). We
use `adapt_delta = 0.95` to reduce divergent transitions in M4.
Convergence is assessed via $\hat{R}$, effective sample size, and the
count of divergent transitions. Models are compared via leave-one-out
cross-validation (Vehtari, Gelman, and Gabry 2017).

# Results

## Convergence

```{r diagnostics}
#| label: tbl-diagnostics
#| tbl-cap: "Convergence summary across the four models."

if (length(models) > 0) {

  track_params   <- c("mu_global", "beta_lmc", "beta_age",
                       "gamma_intercept", "gamma_lmc",
                       "sigma_artist", "sigma_genre",
                       "sigma_lmc_genre", "phi")
  segment_params <- c("mu_global", "beta_chorus", "beta_verse", "beta_sd",
                       "beta_age", "gamma_intercept", "gamma_chorus",
                       "sigma_artist", "sigma_genre",
                       "sigma_chorus_genre", "phi", "chorus_vs_verse")
  timeline_params <- c("mu_global", "beta_lmc", "beta_slope", "beta_curve",
                        "beta_change", "beta_sd", "beta_age",
                        "gamma_intercept", "gamma_lmc",
                        "sigma_genre", "sigma_lmc_genre", "phi")

  param_lists <- list(
    "M1: MuLan Track" = track_params,
    "M2: CLAP Track"  = track_params,
    "M3: Segment"     = segment_params,
    "M4: Timeline"    = timeline_params
  )

  diag_summary <- map2_dfr(models, names(models), function(fit, label) {
    params <- intersect(param_lists[[label]], fit$summary()$variable)
    s <- fit$summary(variables = params)
    tibble(
      model         = label,
      max_rhat      = round(max(s$rhat, na.rm = TRUE), 4),
      min_ess_bulk  = round(min(s$ess_bulk, na.rm = TRUE), 0),
      n_rhat_bad    = sum(s$rhat > 1.01, na.rm = TRUE),
      n_ess_low     = sum(s$ess_bulk < 400, na.rm = TRUE),
      divergences   = sum(fit$diagnostic_summary()$num_divergent)
    )
  })

  kable(diag_summary,
        col.names = c("Model", "Max R̂", "Min ESS bulk",
                      "R̂ > 1.01", "ESS < 400", "Divergences"),
        booktabs = TRUE) %>%
    kable_styling(latex_options = c("hold_position"), full_width = FALSE)
}
```

All four models pass standard convergence diagnostics: maximum
$\hat{R}$ is below 1.003 in every model, minimum bulk ESS exceeds
2,200, and divergent transitions number at most one (in M1) and zero
elsewhere.

## Model Comparison

```{r loo-cv}
#| label: tbl-loo
#| tbl-cap: "Leave-one-out cross-validation comparison. Higher ELPD per song indicates better out-of-sample predictive performance. Sample sizes differ across models because higher-granularity LMC measures (segment, line-by-line) are unavailable for some songs."

if (length(models) > 0) {

  compute_loo_safe <- function(fit, label) {
    ll  <- fit$draws("log_lik", format = "matrix")
    res <- loo(ll, cores = 4)
    N   <- ncol(ll)
    tibble(
      Model     = label,
      N         = N,
      ELPD      = res$estimates["elpd_loo", "Estimate"],
      SE        = res$estimates["elpd_loo", "SE"],
      LOOIC     = res$estimates["looic", "Estimate"],
      ELPD_song = res$estimates["elpd_loo", "Estimate"] / N,
      k_bad     = sum(res$pointwise[, "influence_pareto_k"] > 0.7)
    )
  }

  loo_table <- map2_dfr(models, names(models), compute_loo_safe) %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    arrange(desc(ELPD_song))

  kable(loo_table,
        col.names = c("Model", "N", "ELPD", "SE", "LOOIC",
                      "ELPD/song", "Pareto-k > 0.7"),
        booktabs = TRUE) %>%
    kable_styling(latex_options = c("hold_position"), full_width = FALSE)
}
```

Table \ref{tbl-loo} reports leave-one-out predictive performance. We
focus on ELPD per song, which is comparable across the unequal sample
sizes induced by granularity-driven attrition. The segment model (M3)
achieves the highest predictive density per song at 0.964, edging out
the track-level CLAP model (M2) at 0.948 and the track-level MuLan
model (M1) at 0.947. The trajectory model (M4) trails substantially
at 0.771, largely because its smaller sample provides less information
per parameter. The advantage of M3 over M1 is modest in absolute
terms but consistent with the hypothesis that internal structure of
LMC carries marginal predictive content beyond the track-level
aggregate. The LOO standard errors imply that the ranking of M1, M2,
M3 is not distinguishable from chance at conventional levels.

The marginal advantage of M2 (CLAP) over M1 (MuLan) at the track level
is similarly small. CLAP's broader training distribution may pick up
perceptual features the MuLan playlist-tag prior misses, or the
difference may simply reflect noise. Distinguishing these requires
either a larger corpus or a controlled within-artist comparison.

## Population-Level LMC Effects

```{r lmc-forest}
#| label: fig-lmc-forest
#| fig-cap: "Posterior median and 90% credible interval for the population-level LMC coefficient (logit scale) under each of the four models. M3 reports the chorus coefficient as its primary LMC parameter; M4 reports the mean-LMC coefficient. All credible intervals straddle zero."
#| fig-width: 7
#| fig-height: 3

if (length(models) > 0) {
  lmc_coefs <- bind_rows(
    if (!is.null(fit_m1)) fit_m1$summary("beta_lmc")    %>% mutate(model = "M1: MuLan (track)"),
    if (!is.null(fit_m2)) fit_m2$summary("beta_lmc")    %>% mutate(model = "M2: CLAP (track)"),
    if (!is.null(fit_m3)) fit_m3$summary("beta_chorus") %>% mutate(model = "M3: Segment (chorus)"),
    if (!is.null(fit_m4)) fit_m4$summary("beta_lmc")    %>% mutate(model = "M4: Timeline (mean)")
  )

  ggplot(lmc_coefs, aes(x = median, y = reorder(model, median))) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_errorbarh(aes(xmin = q5, xmax = q95),
                   height = 0.3, linewidth = 1.0, color = "#2A9D8F") +
    geom_point(size = 4, color = "#2A9D8F") +
    labs(x = expression(beta~"(logit scale)"), y = NULL)
}
```

Figure \ref{fig-lmc-forest} reports the population-level LMC
coefficient under each specification. In every case, the posterior
90% credible interval straddles zero: M1 (MuLan track) gives
$\hat{\beta}_{\text{lmc}} = 0.006$ with 90% CI $[-0.131, 0.131]$ and
$P(\beta > 0) = 0.53$; M2 (CLAP) gives median $0.005$ with CI
$[-0.109, 0.122]$; M3's chorus coefficient has median $-0.005$ with
CI $[-0.133, 0.123]$; M4's mean-LMC coefficient has median $-0.034$
with CI $[-0.260, 0.206]$. Hypothesis 1 is therefore *not* supported:
the data neither confirm nor reject a positive population-level LMC
effect at the current sample.

The variance components tell a different story. $\sigma_{\text{artist}}$
is estimated near 0.50 on the logit scale and $\sigma_{\text{genre}}$
near 0.55 — both substantial, reflecting meaningful baseline popularity
heterogeneity at the artist and genre levels. $\sigma_{\text{lmc,
genre}}$, by contrast, is small (around 0.05–0.09), indicating that
the *slope* of LMC on popularity does not vary dramatically across
the four genre clusters. Whether this reflects true homogeneity or
insufficient data to detect heterogeneity cannot be settled here.

## Orientation Moderation

```{r orientation-mod}
#| label: fig-orientation
#| fig-cap: "Posterior densities for the orientation × LMC moderation coefficient $\\gamma_\\text{lmc}$ (or $\\gamma_\\text{chorus}$ for M3) under each model. Positive values indicate stronger LMC–popularity coupling for narrative-forward artists."
#| fig-width: 7
#| fig-height: 3.5

if (length(models) > 0) {
  mod_draws <- bind_rows(
    if (!is.null(fit_m1)) tibble(model = "M1: MuLan Track", gamma = fit_m1$draws(format = "df")$gamma_lmc),
    if (!is.null(fit_m2)) tibble(model = "M2: CLAP Track",  gamma = fit_m2$draws(format = "df")$gamma_lmc),
    if (!is.null(fit_m3)) tibble(model = "M3: Segment",     gamma = fit_m3$draws(format = "df")$gamma_chorus),
    if (!is.null(fit_m4)) tibble(model = "M4: Timeline",    gamma = fit_m4$draws(format = "df")$gamma_lmc)
  )

  ggplot(mod_draws, aes(x = gamma, fill = model)) +
    geom_density(alpha = 0.4) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_fill_brewer(palette = "Set2", name = "Model") +
    labs(x = expression(gamma~"(logit scale)"), y = "Density")
}
```

Figure \ref{fig-orientation} reports the orientation × LMC moderation.
In the track-level models, the moderation coefficient is centered near
zero (M1: $-0.009$, $P(\gamma > 0) = 0.46$; M2: $-0.078$, $P(\gamma >
0) = 0.18$). The segment model gives $\hat{\gamma}_{\text{chorus}} =
0.029$ with $P(\gamma > 0) = 0.64$. The trajectory model produces the
most directional signal: $\hat{\gamma}_{\text{lmc}} = 0.110$ with 90%
CI $[-0.188, 0.394]$ and $P(\gamma > 0) = 0.75$. This pattern — the
moderation effect strengthening as the LMC measure incorporates more
temporal structure — is consistent with the narrative transportation
account, which predicts that fine-grained alignment matters more for
narrative-forward artists. The evidence remains sub-decisive: even in
M4, the credible interval admits a null. Hypothesis 2 is *directionally
supported but inconclusive*.

## Genre Heterogeneity

```{r genre-effects}
#| label: fig-genre
#| fig-cap: "Genre-specific intercepts (top) and genre-varying LMC slopes (bottom) from M1 (track-level MuLan). Genre intercepts are large in magnitude, reflecting baseline popularity differences across clusters. Genre-varying LMC slopes are small and overlapping, reflecting limited differentiation in LMC sensitivity across the four clusters at the current sample size."
#| fig-width: 7
#| fig-height: 6

if (!is.null(fit_m1) && !is.null(master)) {

  n_genre <- length(unique(master$genre_cluster[!is.na(master$genre_cluster) &
                                                 master$genre_cluster != "Other"]))
  genre_levels <- sort(unique(master$genre_cluster[master$genre_cluster != "Other"]))

  genre_int_params <- paste0("alpha_genre[", 1:n_genre, "]")
  genre_int_post <- fit_m1$summary(variables = genre_int_params) %>%
    mutate(genre = genre_levels[as.integer(str_extract(variable, "\\d+"))]) %>%
    filter(!is.na(genre))

  p_genre_int <- ggplot(genre_int_post, aes(x = median, y = reorder(genre, median),
                                             color = genre)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_errorbarh(aes(xmin = q5, xmax = q95), height = 0.3, linewidth = 1.0) +
    geom_point(size = 3.5) +
    scale_color_manual(values = GENRE_COLS, guide = "none") +
    labs(title = "Genre intercepts (baseline popularity)",
         x = expression(alpha[genre]), y = NULL)

  genre_lmc_params <- paste0("beta_lmc_genre[", 1:n_genre, "]")
  genre_lmc_post <- fit_m1$summary(variables = genre_lmc_params) %>%
    mutate(genre = genre_levels[as.integer(str_extract(variable, "\\d+"))]) %>%
    filter(!is.na(genre))

  p_genre_lmc <- ggplot(genre_lmc_post, aes(x = median, y = reorder(genre, median),
                                             color = genre)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_errorbarh(aes(xmin = q5, xmax = q95), height = 0.3, linewidth = 1.0) +
    geom_point(size = 3.5) +
    scale_color_manual(values = GENRE_COLS, guide = "none") +
    labs(title = "Genre-varying LMC slopes",
         x = expression(beta[lmc][","][genre]), y = NULL)

  p_genre_int / p_genre_lmc
}
```

Figure \ref{fig-genre} disaggregates the genre effects in M1. Pop has
the highest baseline popularity, followed by hip-hop, electronic, and
folk/country — consistent with the recency weighting and the heavy
playlist promotion pop receives. The genre-varying LMC slopes are
small with overlapping credible intervals; I cannot claim any one
genre's LMC sensitivity is meaningfully different from another's at
the current sample.

## Segment and Trajectory Decompositions

```{r chorus-vs-verse}
#| label: fig-segment
#| fig-cap: "M3 segment model: posterior densities for chorus LMC, verse LMC, and the chorus–verse contrast. Black bars show 80% and 95% credible intervals. The chorus coefficient is centered slightly above the verse coefficient, but with overlapping intervals."
#| fig-width: 7
#| fig-height: 3.5
#| eval: !expr '!is.null(fit_m3)'

if (!is.null(fit_m3)) {
  m3_draws <- fit_m3$draws(format = "df")

  tibble(
    `Chorus LMC (β_chorus)` = m3_draws$beta_chorus,
    `Verse LMC (β_verse)`   = m3_draws$beta_verse,
    `Chorus − Verse (Δ)`    = m3_draws$chorus_vs_verse
  ) %>%
    pivot_longer(everything(), names_to = "param", values_to = "value") %>%
    ggplot(aes(x = value, y = param, fill = after_stat(x > 0))) +
    stat_halfeye(.width = c(0.8, 0.95), point_interval = median_hdi) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_fill_manual(values = c("FALSE" = "#E76F51", "TRUE" = "#2A9D8F"),
                      guide = "none") +
    labs(subtitle = paste0("P(Chorus > Verse) = ",
                          round(mean(m3_draws$chorus_vs_verse > 0), 3)),
         x = "Coefficient (logit scale)", y = NULL)
}
```

The segment model offers the most directly interpretable internal
structure (Figure \ref{fig-segment}). The chorus coefficient
$\hat{\beta}_{\text{chorus}} = -0.005$ is near zero. The verse
coefficient $\hat{\beta}_{\text{verse}} = -0.035$ with 90% CI
$[-0.091, 0.022]$ is more clearly negative-leaning, with $P(\beta < 0)
= 0.84$. The chorus–verse contrast yields $P(\Delta > 0) = 0.64$ —
directional support for Hypothesis 3a, but short of decisive evidence.

The clearest directional signal in M3 comes from the within-song
consistency parameter: $\hat{\beta}_{\text{sd}} = -0.044$ with 90% CI
$[-0.083, -0.003]$ and $P(\beta_{\text{sd}} < 0) = 0.96$. Because
$\beta_{\text{sd}}$ multiplies the *standard deviation* of segment
LMC, a negative coefficient implies that songs whose LMC is more
consistent across their internal structure are more popular,
controlling for mean LMC. This is the most credible signal in our
current results and provides directional support for Hypothesis 3b.
The magnitude is modest (a one-SD increase in within-song LMC
variability is associated with roughly a 1.1 popularity-point
decrease at the prior mean precision), and the result rests on a
324-song sample.

```{r trajectory}
#| label: fig-trajectory
#| fig-cap: "M4 trajectory model: posterior densities for the five trajectory features of LMC (mean level, slope, curvature, early-to-late change, and consistency)."
#| fig-width: 7
#| fig-height: 4
#| eval: !expr '!is.null(fit_m4)'

if (!is.null(fit_m4)) {
  m4_labels <- c(
    "beta_lmc"    = "Mean LMC (level)",
    "beta_slope"  = "Slope (builds over song?)",
    "beta_curve"  = "Curvature (mid-song peak?)",
    "beta_change" = "Early→Late (strong finish?)",
    "beta_sd"     = "Consistency (stable LMC?)"
  )

  fit_m4$draws(format = "df") %>%
    select(all_of(names(m4_labels))) %>%
    pivot_longer(everything(), names_to = "param", values_to = "value") %>%
    mutate(param = recode(param, !!!m4_labels)) %>%
    ggplot(aes(x = value, y = param, fill = after_stat(x > 0))) +
    stat_halfeye(.width = c(0.8, 0.95), point_interval = median_hdi) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_fill_manual(values = c("FALSE" = "#E76F51", "TRUE" = "#2A9D8F"),
                      guide = "none") +
    labs(x = "Coefficient (logit scale)", y = NULL)
}
```

The trajectory model (Figure \ref{fig-trajectory}) decomposes LMC into
five features. Slope, curvature, and early-to-late change all have
posterior densities centered near zero with wide credible intervals
admitting either sign — Hypothesis 4 is *not* supported at the current
sample. The consistency coefficient $\hat{\beta}_{\text{sd}} = -0.085$
with 90% CI $[-0.172, 0.003]$ and $P(\beta_{\text{sd}} < 0) = 0.94$
replicates the M3 finding on within-song consistency, with wider
intervals reflecting the smaller sample (N = 114). Curvature shows a
posterior median of $-0.068$ with $P(\beta < 0) = 0.90$, mildly
suggestive of an inverted-U over the song's duration but sub-decisive.

## Simulation-Based Parameter Recovery

To validate that the segment model can recover known parameter values
at my observed sample size, I conducted a simulation study. We
generated 50 synthetic datasets calibrated to the observed sample
(N = 350 songs, 30 artists, 4 genres) using a chosen vector of "true"
parameters drawn from reasonable priors (e.g., $\beta_{\text{chorus}}^*
= 0.15$, $\beta_{\text{verse}}^* = 0.05$, $\beta_{\text{sd}}^* = -0.10$).
For each synthetic dataset I re-fit M3 and recorded coverage of the
true value. The model recovers core slope parameters with approximate
calibration: 90% credible intervals cover the truth approximately
80–90% of the time, with mild under-coverage on variance components.
This implies that the credible intervals reported in the empirical
results are roughly calibrated. The exercise does not address
identification under unobserved confounders, only the statistical
coherence of the specification.

# Discussion

The population-level LMC effect on Spotify popularity, controlling
for artist, genre, song age, and audio mood, is statistically
inconclusive at my present sample of 359 songs. Two interpretations
are consistent with the data. Under the first, LMC has no causal
effect on streaming engagement — listeners reward distinctive songs
whether or not their channels are emotionally aligned, and the
theoretical accounts motivating this study are wrong or apply outside
the current corpus. Under the second, LMC has the predicted positive
effect but my power to detect it is insufficient at the current
sample size and data representation. The two interpretations cannot be
distinguished by the present data alone.

What the current evidence *does* support, at directional but
sub-decisive levels of confidence, is a more nuanced picture in which
*structural* features of LMC matter more than its overall level.
Within-song consistency has the most credible signal in both M3 and
M4 (posterior probability of the predicted sign $\geq 0.94$),
suggesting that listeners are more sensitive to the *stability* of
cross-modal alignment than to its average. The chorus–verse contrast
and the orientation × LMC moderation in the trajectory model are
directionally consistent with their respective hypotheses, but
neither rises to confirmatory evidence. Separately, the segment model
(M3) offers the best out-of-sample fit — consistent with a story in
which the internal structure of LMC carries information the
song-level aggregate discards.

Several threats to inference apply. First, the data-generating
process described in Section \ref{conceptual-framework-and-hypotheses}
implies unblockable backdoor paths from artist and genre into
popularity through unobserved promotion, playlist placement, and
social distribution. I have no credible instrument for LMC at the
song level. The strongest available identification is *within-artist*
variation, but with roughly 10 songs per artist this is thin. Second,
our controls absorb observed audio characteristics but not
listener-level effects driving streaming behavior at scale. Third,
the outcome itself is the product of Spotify's algorithm; songs more
aligned with playlist contexts may receive more algorithmic exposure,
biasing the LMC–popularity association in either direction depending
on what the algorithm rewards. These threats place a ceiling on the
inferential force of any single-cross-section analysis with
publicly-scraped data. The path forward is therefore as much about
broadening data and tightening identification as fitting better
models.

# Limitations and the Path Forward

The structure of the next phase follows directly from the limitations
identified above. I organize it under four priorities.

## Sample Expansion and Embedding Robustness

The most binding constraint is sample size. my 359 songs deliver
insufficient power to distinguish a small-but-real LMC effect from a
true null. I will expand the corpus to additional artists within the
four existing clusters and to additional genres (R&B, Latin pop,
indie rock, classical, and jazz where relevant) so that the
production–narrative spectrum is more densely populated, and move from
purposive artist sampling toward a *systematic sample from Spotify's
recommendation distribution* — seeding the algorithm with anchor
tracks across the popularity range and crawling the neighborhood of
each anchor, yielding a sample that better represents the full
popularity distribution rather than concentrating in the right tail.
In parallel, I will complete the MERT + Sentence-BERT late-fusion
baseline and re-estimate all four models: comparable predictive
validity would imply joint training contributes little, while inferior
performance would directly evidence the value of contrastive joint
training. I will also run M3 and M4 on CLAP-derived segment and
line-level LMC; at present these higher-granularity analyses use only
MuLan.

## Refining Segmenting and Trajectory Analysis

Our higher-granularity measures suffer from sparsity (M3: 324/359;
M4: 114/359) driven by failures of WhisperX forced alignment. Planned
improvements include structurally-aware segmentation that uses lyric
section headers as anchors and falls back to alignment only within
sections; alternative aligners (e.g., NeMo's ASR models, Koluguri,
Park, and Ginsburg 2022); and an examination of whether segment
*ordering* matters — does the sequence of high- and low-LMC segments
predict popularity beyond their average and variance? — as a richer
test of Hypothesis 4.

## Generation Experiments and Human Validation

The strongest evidence for a causal LMC effect would come from
experimental manipulation. I plan a generation experiment in which
a base song is counterfactually re-lyriced with content of varying
emotional alignment to its musical bed, holding other features fixed.
Generative lyric models conditioned on a target embedding region,
combined with controlled human evaluation, would isolate the causal
effect of LMC on listener-level outcomes (liking, predicted listening
time, repeat-listening intent) free from observational confounds.
This track also produces *human reference scores* providing external
validation of the cosine similarity itself — a machine-internal
quantity that should be calibrated against human judgment before
being treated as a marketing-relevant construct.

## Better Outcomes and Sharper Identification

Our coarsest constraint is the outcome variable. Spotify's popularity
index aggregates discovery, retention, repeat listening, skip
behavior, and shareability into a single bounded score, each
plausibly affected differently by LMC: transportation predicts higher
retention and repeat-listening rates, while fluency predicts higher
initial liking. Discriminating these requires listener-level
behavioral data of a kind no public API provides; I will continue
to pursue a research partnership with Spotify, Apple Music, or
another platform. In its absence I will broaden outcomes using
public signals — chart position, playlist appearances, and per-song
streams from financial filings.

Identification will improve from two further sources of within-unit
variation. *Within-album* comparisons restrict to songs released
simultaneously on the same album, absorbing essentially all
album-launch promotional shocks. *Cover versions* — multiple
recordings of the same composition by different artists — hold
lyrical content fixed while varying audio production, so that LMC
variation is by construction attributable to production choices
rather than lyric writing.

\bigskip

We close by noting what this paper does and does not claim. It does
not claim to have established that lyric–music congruence raises
streaming popularity. It does claim to have built a measurement
pipeline operationalizing a previously elusive construct, specified
a hierarchical Bayesian framework appropriate to the data structure,
produced an honest first-pass empirical estimate, and laid out the
design choices, identification threats, and next steps that will
determine whether the construct matters commercially.

# References {.unnumbered}

\begingroup
\setlength{\parindent}{-0.4in}
\setlength{\leftskip}{0.4in}
\noindent

Areni, C. S., and D. Kim. 1993. "The Influence of Background Music on
Shopping Behavior: Classical versus Top-Forty Music in a Wine Store."
*Advances in Consumer Research* 20: 336–340.

Askin, N., and M. Mauskapf. 2017. "What Makes Popular Culture Popular?
Product Features and Optimal Differentiation in Music." *American
Sociological Review* 82(5): 910–944.

Bain, M., J. Huh, T. Han, and A. Zisserman. 2023. "WhisperX: Time-Accurate
Speech Transcription of Long-Form Audio." *Proceedings of Interspeech*
2023: 4489–4493.

Berger, J., A. Humphreys, S. Ludwig, W. W. Moe, O. Netzer, and D. A.
Schweidel. 2020. "Uniting the Tribes: Using Text for Marketing Insight."
*Journal of Marketing* 84(1): 1–25.

Berlyne, D. E. 1971. *Aesthetics and Psychobiology*. New York:
Appleton-Century-Crofts.

Bogdanov, D., N. Wack, E. Gómez, S. Gulati, P. Herrera, O. Mayor, G.
Roma, J. Salamon, J. R. Zapata, and X. Serra. 2013. "Essentia: An Audio
Analysis Library for Music Information Retrieval." *Proceedings of the
14th International Society for Music Information Retrieval Conference
(ISMIR)*: 493–498.

Dew, R., and A. Ansari. 2018. "Bayesian Nonparametric Customer Base
Analysis with Model-Based Visualizations." *Marketing Science* 37(2):
216–235.

Eerola, T., and J. K. Vuoskoski. 2011. "A Comparison of the Discrete and
Dimensional Models of Emotion in Music." *Psychology of Music* 39(1):
18–49.

Ferrari, S., and F. Cribari-Neto. 2004. "Beta Regression for Modelling
Rates and Proportions." *Journal of Applied Statistics* 31(7): 799–815.

Fong, H., V. Kumar, and K. Sudhir. 2024. "A Theory-Based Explainable Deep
Learning Architecture for Music Emotion." Working paper, Yale School of
Management. arXiv:2408.07113.

Gabrielsson, A. 2002. "Emotion Perceived and Emotion Felt: Same or
Different?" *Musicae Scientiae* 5(1, supplement): 123–147.

Green, M. C., and T. C. Brock. 2000. "The Role of Transportation in the
Persuasiveness of Public Narratives." *Journal of Personality and Social
Psychology* 79(5): 701–721.

Hekkert, P. 2006. "Design Aesthetics: Principles of Pleasure in Design."
*Psychology Science* 48(2): 157–172.

Huang, Q., A. Jansen, J. Lee, R. Ganti, J. Y. Li, and D. P. W. Ellis.
2022. "MuLan: A Joint Embedding of Music Audio and Natural Language."
*Proceedings of the 23rd International Society for Music Information
Retrieval Conference (ISMIR)*: 277–285.

IFPI. 2024. *Global Music Report 2024*. London: International Federation
of the Phonographic Industry.

Koluguri, N. R., T. Park, and B. Ginsburg. 2022. "TitaNet: Neural Model
for Speaker Representation with 1D Depth-Wise Separable Convolutions and
Global Context." *Proceedings of the IEEE International Conference on
Acoustics, Speech and Signal Processing (ICASSP)*: 8102–8106.

Landwehr, J. R., A. A. Labroo, and A. Herrmann. 2011. "Gut Liking for
the Ordinary: Incorporating Design Fluency Improves Automobile Sales
Forecasts." *Marketing Science* 30(3): 416–429.

Lee, A. Y., and A. A. Labroo. 2004. "The Effect of Conceptual and
Perceptual Fluency on Brand Evaluation." *Journal of Marketing Research*
41(2): 151–165.

Lena, J. C., and R. A. Peterson. 2008. "Classification as Culture: Types
and Trajectories of Music Genres." *American Sociological Review* 73(5):
697–718.

Ma, Y., R. Yuan, Y. Li, G. Zhang, X. Chen, H. Yin, C. Lin, A. Ragni, E.
Benetos, N. Gyenge, et al. 2023. "MERT: Acoustic Music Understanding
Model with Large-Scale Self-Supervised Training." arXiv:2306.00107.

MacInnis, D. J., and C. W. Park. 1991. "The Differential Role of
Characteristics of Music on High- and Low-Involvement Consumers'
Processing of Ads." *Journal of Consumer Research* 18(2): 161–173.

Melvill-Smith, S., A. E. Krause, and A. C. North. 2023. "Song Popularity
and Processing Fluency of Lyrics." *Psychology of Music* 51(3): 712–726.

North, A. C., and D. J. Hargreaves. 1995. "Subjective Complexity,
Familiarity, and Liking for Popular Music." *Psychomusicology: A Journal
of Research in Music Cognition* 14(1–2): 77–93.

North, A. C., D. J. Hargreaves, and J. McKendrick. 1999. "The Influence
of In-Store Music on Wine Selections." *Journal of Applied Psychology*
84(2): 271–276.

Ordanini, A., J. C. Nunes, and A. Nanni. 2018. "The Featuring Phenomenon
in Music: How Combining Artists of Different Genres Increases a Song's
Popularity." *Marketing Letters* 29(4): 485–499.

Panda, R., R. Malheiro, and R. P. Paiva. 2023. "Audio Features for Music
Emotion Recognition: A Survey." *IEEE Transactions on Affective
Computing* 14(1): 68–88.

Phillips, B. J., and E. F. McQuarrie. 2010. "Narrative and Persuasion
in Fashion Advertising." *Journal of Consumer Research* 37(3): 368–392.

Radford, A., J. W. Kim, C. Hallacy, A. Ramesh, G. Goh, S. Agarwal, G.
Sastry, A. Askell, P. Mishkin, J. Clark, G. Krueger, and I. Sutskever.
2021. "Learning Transferable Visual Models from Natural Language
Supervision." *Proceedings of the 38th International Conference on
Machine Learning (ICML)*: 8748–8763.

Reber, R., N. Schwarz, and P. Winkielman. 2004. "Processing Fluency and
Aesthetic Pleasure: Is Beauty in the Perceiver's Processing Experience?"
*Personality and Social Psychology Review* 8(4): 364–382.

Reimers, N., and I. Gurevych. 2019. "Sentence-BERT: Sentence Embeddings
Using Siamese BERT-Networks." *Proceedings of the 2019 Conference on
Empirical Methods in Natural Language Processing (EMNLP)*: 3982–3992.

Salganik, M. J., P. S. Dodds, and D. J. Watts. 2006. "Experimental Study
of Inequality and Unpredictability in an Artificial Cultural Market."
*Science* 311(5762): 854–856.

Smithson, M., and J. Verkuilen. 2006. "A Better Lemon Squeezer?
Maximum-Likelihood Regression with Beta-Distributed Dependent Variables."
*Psychological Methods* 11(1): 54–71.

Spotify AB. 2024. "Spotify Web API Reference: Track Object — Popularity."
https://developer.spotify.com/documentation/web-api/reference/get-track.
Accessed April 2024.

Stan Development Team. 2024. *cmdstanr: R Interface to CmdStan*. R
package version 0.7.

Toubia, O., G. Iyengar, R. Bunnell, and A. Lemaire. 2019. "Extracting
Features of Entertainment Products: A Large-Scale Simultaneous
Segmentation and Estimation Approach." *Marketing Science* 38(1):
187–207.

Van Laer, T., K. De Ruyter, L. M. Visconti, and M. Wetzels. 2014. "The
Extended Transportation-Imagery Model: A Meta-Analysis of the
Antecedents and Consequences of Consumers' Narrative Transportation."
*Journal of Consumer Research* 40(5): 797–817.

Vehtari, A., A. Gelman, and J. Gabry. 2017. "Practical Bayesian Model
Evaluation Using Leave-One-Out Cross-Validation and WAIC." *Statistics
and Computing* 27(5): 1413–1432.

Whitman, B. 2005. *Learning the Meaning of Music*. PhD dissertation,
Massachusetts Institute of Technology.

Winkielman, P., J. Halberstadt, T. Fazendeiro, and S. Catty. 2006.
"Prototypes Are Attractive Because They Are Easy on the Mind."
*Psychological Science* 17(9): 799–806.

Wu, Y., K. Chen, T. Zhang, Y. Hui, T. Berg-Kirkpatrick, and S. Dubnov.
2023. "Large-Scale Contrastive Language–Audio Pretraining with Feature
Fusion and Keyword-to-Caption Augmentation." *Proceedings of the IEEE
International Conference on Acoustics, Speech and Signal Processing
(ICASSP)*: 1–5.

Yang, Y.-H., and H. H. Chen. 2012. "Machine Recognition of Music Emotion:
A Review." *ACM Transactions on Intelligent Systems and Technology* 3(3):
40:1–40:30.

Zajonc, R. B. 1968. "Attitudinal Effects of Mere Exposure." *Journal of
Personality and Social Psychology Monograph Supplement* 9(2, Pt. 2): 1–27.

\endgroup
