---
title: "Lyric–Music Congruence → Streaming Popularity"
subtitle: "Embedding sweep — 4 models × 3 prompts × 5 structures (60 fits, v4 models)"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    fig-width: 9
    fig-height: 5
    theme: cosmo
    embed-resources: true
params:
  controls: "mert"
execute:
  warning: false
  message: false
  echo: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages(source(here::here("sweep", "R", "report_helpers_sweep.R")))
library(knitr)
CTRL <- params$controls

loo_tab   <- load_loo_sweep()
corpus_df <- tryCatch(build_corpus_sweep(controls = CTRL)$df, error = function(e) NULL)
N_corpus  <- if (!is.null(corpus_df)) nrow(corpus_df) else NA_integer_
genre_names <- if (!is.null(corpus_df)) levels(factor(corpus_df$genre)) else NULL

# Load every fit named in the LOO table (skips any missing .rds). extract_fit GCs
# each fit after summarising, so all 60 can be read sequentially.
tags <- loo_tab$tag[file.exists(file.path(OUTPUT, paste0(loo_tab$tag, ".rds")))]
M <- setNames(lapply(tags, extract_fit, corpus_df = corpus_df, genre_names = genre_names), tags)
M <- Filter(Negate(is.null), M)
diag_tbl <- bind_rows(lapply(M, function(m) m$diag))

emb_split <- function(e) { p <- strsplit(e, "_", fixed = TRUE)[[1]]; list(model = p[1], prompt = p[2]) }
```

> **What this is.** The full **model × prompt × structure sweep**: 4 embedding models
> (MuLan, LAION-CLAP, MS-CLAP, CLaMP 3) × 3 text prompts (raw / contains / idea) = 12
> embeddings, each fit at 5 structures (song-wide + line ±1/±5/±10 s as track models,
> chorus/verse as a segment model) = **60 hierarchical Beta-regression fits**, all on
> **one shared complete-case corpus of `r N_corpus` songs** so every LOO is directly
> comparable. Controls: **`r CTRL`**. All effects are on the **logit scale**.

# Executive summary

```{r exec}
#| results: asis
if (length(M) == 0) {
  cat("_No fits found in `sweep/output/`. Run `Rscript sweep/R/run_models_sweep.R` first._")
} else {
  mg <- loo_marginals(loo_tab)
  best <- loo_tab %>% slice_max(elpd_loo, n = 1)
  cat(sprintf("
- **Fits:** %d of 60 present, all on the same **%s** songs (LOO directly comparable).
- **Sampler health:** min E-BFMI **%.2f**, max R-hat **%.3f**, **%d** total divergences, worst Pareto-k count **%d** (across %d fits).
- **Best single fit (LOO):** **%s** (%s · %s · %s), elpd **%.1f**.
- **Best model (mean elpd):** **%s**. **Best prompt:** **%s**. **Best structure:** **%s**.
- Read §2 for the full 60-fit ranking; §3 for the LMC effect sizes; §5 for posterior checks.
",
  length(M), N_corpus, min(diag_tbl$ebfmi_min), max(diag_tbl$rhat_max), sum(diag_tbl$n_div),
  max(loo_tab$k_gt_0.7), nrow(diag_tbl),
  best$tag, MODEL_LABEL[[best$model]], best$prompt, STRUCTURE_LABEL[[as.character(best$structure)]], best$elpd_loo,
  MODEL_LABEL[[as.character(mg$by_model$model[1])]], mg$by_prompt$prompt[1],
  STRUCTURE_LABEL[[as.character(mg$by_structure$structure[1])]]))
}
```

# 1 · Sampler health (all fits)

```{r diag}
diag_tbl %>%
  mutate(model = MODEL_LABEL[map_chr(embedding, ~emb_split(.x)$model)],
         prompt = map_chr(embedding, ~emb_split(.x)$prompt),
         Structure = ifelse(is.na(measure), "segment", paste0("track_", measure))) %>%
  transmute(Model = model, Prompt = prompt,
            Structure = STRUCTURE_LABEL[Structure], Divergences = n_div,
            `Min E-BFMI` = round(ebfmi_min, 2), `Max R-hat` = round(rhat_max, 3),
            `Min bulk ESS` = round(ess_bulk_min)) %>%
  arrange(desc(Divergences), desc(`Max R-hat`)) %>%
  head(15) %>%
  kable(align = "lllrrrr", caption = "15 least-healthy fits (all 60 in loo_all_sweep.csv). Watch divergences > 0, R-hat > 1.01, ESS < 400.")
```

# 2 · Model comparison (LOO) — the whole grid

All fits share the same `r N_corpus` songs, so a single LOO ranking is valid across
**every** model × prompt × structure.

```{r loo-heatmap, fig.height=6}
#| fig-cap: "LOO Δelpd across all 60 fits (0 = best; more negative = worse predictively)."
plot_loo_heatmap(loo_tab)
```

```{r loo-top}
loo_tab %>%
  transmute(Rank = row_number(), Model = MODEL_LABEL[model], Prompt = prompt,
            Structure = STRUCTURE_LABEL[as.character(structure)],
            elpd = round(elpd_loo, 1), SE = round(se, 1),
            `p_loo` = round(p_loo, 1), `k>0.7` = k_gt_0.7) %>%
  head(15) %>%
  kable(align = "rlllrrrr", caption = "Top 15 fits by elpd_loo (full table: sweep/output/loo_all_sweep.csv).")
```

```{r loo-marginals}
mg <- loo_marginals(loo_tab)
knitr::kables(list(
  kable(mg$by_model %>% transmute(Model = MODEL_LABEL[as.character(model)], `Mean elpd` = round(mean_elpd, 1)), align = "lr"),
  kable(mg$by_prompt %>% transmute(Prompt = as.character(prompt), `Mean elpd` = round(mean_elpd, 1)), align = "lr"),
  kable(mg$by_structure %>% transmute(Structure = STRUCTURE_LABEL[as.character(structure)], `Mean elpd` = round(mean_elpd, 1)), align = "lr")
), caption = "Marginal mean elpd — which model, which prompt, which structure predicts best on average.")
```

```{r loo-compare-note}
#| results: asis
cat("_`Δelpd` between two fits is credible only if it exceeds ~2× its SE. The full pairwise ",
    "`loo_compare` object is saved to `sweep/output/loo_compare_all_sweep.rds`._")
```

# 3 · How big is the LMC effect? (by structure, model, prompt)

Track-model LMC slope `beta_lmc` (song-wide + line windows), and the segment
chorus−non-chorus contrast, across the grid.

```{r beta-lmc, fig.height=6}
#| fig-cap: "Track-model LMC slope (median ± 90% CI), faceted by structure; colour = model, shape = prompt."
track_tags <- names(M)[startsWith(names(M), "track_")]
blmc <- map_dfr(track_tags, function(tg) {
  m <- M[[tg]]; s <- m$summ %>% filter(param == "beta_lmc")
  if (!nrow(s)) return(NULL)
  sp <- emb_split(m$embedding)
  tibble(structure = paste0("track_", m$measure), model = sp$model, prompt = sp$prompt,
         median = s$median, q05 = s$q05, q95 = s$q95, p_pos = s$p_pos)
})
blmc <- blmc %>% mutate(structure = factor(STRUCTURE_LABEL[structure], levels = STRUCTURE_LABEL[STRUCTURE_LEVELS]),
                        model = MODEL_LABEL[model])
ggplot(blmc, aes(median, model, colour = model, shape = prompt)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  geom_pointrange(aes(xmin = q05, xmax = q95), position = position_dodge(width = 0.6)) +
  facet_wrap(~ structure, ncol = 2) +
  labs(x = "LMC slope on popularity (logit)", y = NULL, colour = NULL, shape = "prompt") +
  theme_sweep() + theme(legend.position = "bottom") + guides(colour = "none")
```

```{r seg-delta, fig.height=4.5}
#| fig-cap: "Segment models: chorus − non-chorus congruence slope (Δ), by model and prompt."
seg_tags <- names(M)[startsWith(names(M), "segment_")]
segd <- map_dfr(seg_tags, function(tg) {
  m <- M[[tg]]; s <- m$summ %>% filter(param == "chorus_vs_nonchorus")
  if (!nrow(s)) return(NULL)
  sp <- emb_split(m$embedding)
  tibble(model = MODEL_LABEL[sp$model], prompt = sp$prompt,
         median = s$median, q05 = s$q05, q95 = s$q95)
})
ggplot(segd, aes(median, model, colour = prompt)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  geom_pointrange(aes(xmin = q05, xmax = q95), position = position_dodge(width = 0.5)) +
  labs(x = "Chorus − non-chorus slope (logit; <0 ⇒ verse congruence predicts better)",
       y = NULL, colour = "prompt") + theme_sweep() + theme(legend.position = "bottom")
```

# 4 · Best-fit spotlight

The single top-LOO fit, in detail: population effects + control block, and the
genre-varying LMC slope.

```{r spotlight, fig.height=4.5}
#| fig-cap: "Population effects and control block for the best-LOO fit (median, 90% CI)."
best_tag <- loo_tab$tag[which.max(loo_tab$elpd_loo)]
mb <- M[[best_tag]]
if (!is.null(mb)) plot_effects(mb)
```

```{r spotlight-genre, fig.height=4.5}
#| fig-cap: "Genre-varying congruence slope for the best-LOO fit (violin = posterior, point = median ± 90%)."
if (!is.null(mb)) plot_genre_slope(mb)
```

# 5 · Posterior predictive checks

Observed popularity vs. posterior-replicated draws. To stay legible across 60 fits
we show the **best fit per structure** (swap `pick` below to inspect others).

```{r ppc, fig.height=6}
#| fig-cap: "Observed (dark) vs posterior-predicted (light) popularity — best fit per structure."
pick <- loo_tab %>% group_by(structure) %>% slice_max(elpd_loo, n = 1) %>% ungroup()
ppc_M <- Filter(function(m) !is.null(m$ppc), M[pick$tag])
if (length(ppc_M)) {
  ppc_df <- bind_rows(lapply(ppc_M, function(m) {
    obs <- tibble(value = m$ppc$y, kind = "observed", draw = 0L)
    rep <- as_tibble(t(m$ppc$yrep)) %>% mutate(.id = row_number()) %>%
      pivot_longer(-.id, names_to = "draw", values_to = "value") %>%
      mutate(kind = "replicated", draw = as.integer(factor(draw)))
    sp <- emb_split(m$embedding)
    lbl <- sprintf("%s\n%s · %s · %s", m$tag, MODEL_LABEL[[sp$model]], sp$prompt,
                   STRUCTURE_LABEL[[if (is.na(m$measure)) "segment" else paste0("track_", m$measure)]])
    bind_rows(obs, rep) %>% mutate(panel = lbl)
  }))
  ggplot(ppc_df) +
    geom_density(data = ~filter(.x, kind == "replicated"),
                 aes(value, group = draw), colour = "#c9d7ee", linewidth = 0.15) +
    geom_density(data = ~filter(.x, kind == "observed"), aes(value), colour = "#1f6feb", linewidth = 0.9) +
    facet_wrap(~ panel, ncol = 2) +
    labs(x = "Spotify popularity / 100 (boundary-adjusted)", y = NULL) + theme_sweep()
} else cat("_No PPC available — ensure the shared corpus rebuilt (build_corpus_sweep)._")
```

# 6 · Notes & next steps

```{r next}
#| results: asis
cat(sprintf("
- **§2 is the headline:** the 60-fit LOO grid on the same %s songs. Two fits are only
  credibly different when Δelpd exceeds ~2× its SE.
- **Marginals (§2):** which *model*, *prompt*, and *structure* win on average — the
  cleanest read on whether MS-CLAP / CLaMP 3 or the caption-style prompts help.
- **§3:** effect sizes; watch whether the sign/size of the LMC slope is stable across
  prompts within a model (prompt-sensitivity is itself a finding).
- **Corpus size:** the global complete-case rule means any CLaMP 3 / MS-CLAP missingness
  shrinks N for *all* fits — check the %s-song count against the per-cell coverage in the
  notebook's `build_master.missing_report()`.
- Controls here are **%s**; re-run the notebook + `run_models_sweep.R` with a different
  control set to test robustness.
", N_corpus, N_corpus, CTRL))
```

---

*Rendered from `sweep/lmc_report_sweep.qmd`. Fit the battery with
`Rscript sweep/R/run_models_sweep.R` (4 chains × 1000+1000, MERT controls).*
